---
title: "Predictors of Opioid Misuse Disorder in the US"
subtitle: "Final Project"
format: pdf
bibliography: references.bib
csl: apa.csl
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong
  - name: Said Naqwe
date: April 26, 2026
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Nana Frimpong, Said Naqwe}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: load-packages
#| include: false
# Load necessary libraries
library(tidyverse)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
library(pROC)
library(tidytable)
library(stringr)
library(conflicted)
library(caret)
library(forecast)
library(glmnet)
library(MASS)

# Specify package preferences to avoid function name conflicts
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::group_by)
conflicts_prefer(dplyr::ungroup)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::n_distinct)
conflicts_prefer(dplyr::full_join)
conflicts_prefer(dplyr::bind_rows)
conflicts_prefer(dplyr::bind_cols)
conflicts_prefer(dplyr::rename)
conflicts_prefer(dplyr::pull)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::transmute)
conflicts_prefer(tidyr::pivot_wider)
conflicts_prefer(tidyr::pivot_longer)
```

::: callout-note
From Alex: Some sections have our names next to the section title and others don't. I'm fine with either, but could we make them consistent?
:::

# Abstract

::: callout-note
**Naomi to complete.** Use 250 words or less.
:::

Background: Opioid addiction is a public health crisis that has affected countless lives.

Research Question: What factors are important for predicting prevalence rate of opioid use disorder (OUD)?

Methods: We downloaded data on the prevalence of pain reliever misuse in each state from the National Survey on Drug Use and Health from the Substance and Mental Health Services Administration for all available years (2016–2019 and 2022)...\[TBD\]

Results: \[TBD\]

Conclusion: \[TBD\]

# Key words

Opioid Use Disorder, Regression Analysis, Public Health, Substance Use Disorder

# Introduction

Opioid addiction is a public health crisis that has affected countless lives. The United States is experiencing a public health crisis of almost unprecedented scale: an epidemic of opioid use disorder (OUD) and related overdose deaths, the magnitude of which has increased exponentially in the past decades [@Mancher2019]. In just one year, the age-adjusted rate of drug overdose deaths increased 31% (from 21.6 in 2019 per 100k population to 28.3 in 2020) [@Hedegaard2021]. In this project, we investigate what factors are important for predicting the prevalence rate of OUD across US states.

# Literature review

OUD is a complex problem with many risk factors, including sociodemographic, geographic, and community/societal factors. [@Jalali2020] Sociodemographic risk factors include age, sex, and occupation. Opioid misuse is most common in young adults (18-25 years old), and first misuse of opioids at an early age is a significant risk factor for OUD. [@Crane2013][@Schepis2017] In addition, male sex is a risk factor for OUD because men are more likely than women to experience euphoria from opioids. [@Corner2010] Some occupations, particularly those that involve heavy physical labor (e.g., construction, mining), are also risk factors for OUD, because they may be associated with more work-related injuries that lead to chronic pain and treatment with opioid analgesics. [@Shaw2020][@Friedman2025] In addition to these sociodemographic factors, mental health conditions, such as anxiety and depression, and a history of misusing other subtances (eg, illicit drugs) can increase the risk of OUD. [@Valente2020][@Cicero2020]

The risk of OUD also varies by geographic location. For example, opioid medications are prescribed more frequently in non-metropolitan areas, which may be due to the higher proportion of older individuals in these areas who need pain relief after surgeries or individuals who are injured while performing manual labor. [@Keyes2014][@Rigg2015] Rural areas may also have fewer community resources for prevention and treatment programs than urban areas, which can increase the likelihood of developing and prolonging OUD. [@EDC2016]

Furthermore, disparities in healthcare access and insurance coverage can make it more difficult for some individuals to receive care for OUD, particularly those who are racial or ethnic minorities, live in rural areas, or cannot afford health insurance. [@Latimore2021][@PEW2019] The US Affordable Care Act (ACA; also known as “Obamacare”) provided states the option to extend Medicaid benefits to non-elderly adults with lower incomes, including access to OUD treatment; however, not all states have implemented the expansion. [@Mitchell2021] These differences in the implementation of ACA have resulted in a lower rate of prescriptions for medications to treat OUD in states that have not yet expanded Medicaid compared with those that have. [@Sharp2018][@Lewis2020]

Although many studies have documented risk factors for OUD in various groups (e.g., military veterans) or geographic regions (e.g, urban verus rural), few have analyzed nationwide data. One study, which modeled National Survey on Drug Use and Health (NSDUH) data from 2017-2018 using multivariate logistic regression, found that the most significant predictors of OUD were abuse of other substances, particularly tranquilizers (odds ratio \[OR\]=16.7), criminal history (OR=2.6), and lack of private health insurance (OR=2.1) (all P\<0.001). [@Montiel_Ishino2020] Here, we perform a similar analysis using a broader range of NSDUH data (2016-2019 + 2022), including data before and after the Covid pandemic.

# Methodology

::: callout-note
Discuss the key aspects of your problem, data set and regression model(s). Given that you are working on real-world data, explain at a high-level your exploratory data analysis, how you prepared the data for regression modeling, your process for building regression models, and your model selection.
:::

We use data on the state-level prevalence of pain reliever misuse and demographic characteristics from the National Survey on Drug Use and Health (NSDUH) from the Substance and Mental Health Services Administration for 2018-2024 [@NSDUH2018-2019].

## Exploratory Data Analysis

We performed exploratory data analysis to understand the distribution of our outcome variable (prevalence of pain reliever misuse) and the relationships between our predictors and the outcome. We visualized the correlations between predictors and the outcome variable, as well as the distribution of the numeric variables.

## Data Preparation (Alex)

Box-Cox transformations were used to normalize data distributions of numeric predictors. We confirmed that these variables had positive values, which is required for this method. We explicitly calculated the optimal lambda values, which provided insight into the optimal transformations needed to normalize each variable, and then used these values to calculate the transformed data.

The data did not contain any missing values, so imputation was not needed. For consistency with a similar study [@Montiel_Ishino2020] (see Literature review section), which will be used for comparison in the Discussion section, no additional features were created.

Finally, we split the data 70:30 into training and test sets.

## Assumptions (Nana)

## Build models methods (Richie)

For our models, we will attempt to predict the state-level prevalence of pain reliever misuse by evaluating a series of regression techniques designed to balance model fit with predictive accuracy. Our methodology follows a multi-stage approach, transitioning from a comprehensive baseline to restricted models that handle the complexities of high-dimensional demographic data.  

### Baseline Full Multiple Linear Regression
Our initial approach involves constructing a Full Multiple Linear Regression model using all 20+ available predictors. This process partitions the response variable into a systematic component represented by the predictors and a random error component. We use the method of Least Squares to calculate the unique set of coefficients that minimizes the sum of squared residuals, ensuring we obtain the best linear unbiased estimates (BLUE) given our data. To execute this, we will fit a model using every numeric demographic and socioeconomic variable in the training set to establish a performance ceiling for our $R^2$ and residual standard error.

### Stepwise Variable Selection

To address potential redundancy in our 20+ predictors, we implement Stepwise Selection based on the Akaike Information Criterion (AIC). This "auto-selection" method iteratively adds or removes variables, comparing a larger model ($\Omega$) to a smaller subset ($\omega$). The goal is to minimize the AIC, which effectively penalizes the model for every additional parameter included, thereby favoring simpler, more interpretable explanations.

### LASSO Regression
Because demographic variables (e.g., age, race, and poverty levels) are often highly correlated, we will utilize LASSO (Least Absolute Shrinkage and Selection Operator). This technique introduces a penalty term to the OLS equation that can shrink insignificant coefficients exactly to zero. This process requires Hyperparameter Tuning to find the optimal value of $\lambda$ (the penalty strength) that produces the minimum error. 

### Predictive Validation and Testing
To ensure our models generalize to unseen data, we utilize a two-stage validation process. First, we will implement 10-fold Cross-Validation on the training data. This involves splitting the training set into ten equal parts, or "folds"; the model is trained on nine folds and validated on the tenth, with the process repeating ten times so every data point is used for validation once. We use the average Root Mean Squared Error (RMSE) across these folds to select our winning model. Finally, we will test the winning model on the held-out test set to get an honest estimate of its predictive ability on new data.

## Select models methods (Said)

# Experimentation and Results

::: callout-note
Describe the specifics of what you did (data exploration, data preparation, model building, model selection, model evaluation, etc.), and what you found out (statistical analyses, interpretation and discussion of the results, etc.).
:::

```{r}
#| label: load-join-data

# Read in all CSVs from the data directory and store them in a list
filepaths <- list.files(path = "data", pattern = "*.csv", full.names = TRUE)

# Helper to extract the "YY-YY" year substring from filenames
extract_year_tag <- function(path) {
  fname <- basename(path)
  tag <- stringr::str_extract(fname, "\\d{2}-\\d{2}")
  if (is.na(tag)) tag <- gsub("\\.csv$", "", fname)
  tag
}

# Read each file and add a 'year' column based on the filename (e.g. "18-19")
data_list <- lapply(filepaths, function(fp) {
  df <- readr::read_csv(fp, show_col_types = FALSE) |> 
    janitor::clean_names()
  
  # Get name of first column (variable of interest)
  first_col <- names(df)[1]

  # Transform each df.
  df <- df |> 
    # Filter out redundant ("Overall") rows
    filter(
      state_us_abbreviation != "Overall") |>

    # Calculate precise percentages
    group_by(state_us_abbreviation) |>
    mutate(
      denominator = weighted_count[column_percent == 1]
    ) |>
    ungroup() |> 
    mutate(
      percent = weighted_count / denominator
    ) |> 

    # Get subset of columns
    select(1, state_us_abbreviation, percent) |> 

    # Pivot first column wider
    mutate(
      "{first_col}" := paste0(first_col, "_", as.character(.data[[first_col]]))
    ) |>
    pivot_wider(names_from = 1, values_from = percent) |> 

    # Drop columns with all the same values
    select(where(~ n_distinct(.) > 1)) |>

    # Add year column
    mutate(
      year = extract_year_tag(fp)
    ) |>

    # Clean names
    janitor::clean_names()

  df
})

# Get list of data years
years <- sapply(filepaths, extract_year_tag) |> unique()

# Iterate across years, join data frames in data_list by state and year within each year.
yearly_joined_data <- lapply(years, function(yr) {
  year_dfs <- data_list |>
    purrr::keep(~ n_distinct(.x$year) == 1 && .x$year[[1]] == yr)

  if (length(year_dfs) == 0) return(NULL)
  if (length(year_dfs) == 1) return(year_dfs[[1]])

  Reduce(function(x, y) full_join(x, y, by = c("state_us_abbreviation", "year")), year_dfs)
})

# Append yearly_joined_data across all years into a single data frame
# Filter out NULL values first to avoid bind_rows errors
df <- yearly_joined_data |> 
  purrr::discard(is.null) |>
  bind_rows()
```

```{r}
#| label: rename-columns

df_renamed <- df |> 
  rename(
    state = state_us_abbreviation,

    age_12_17 = rc_adult_age_category_recode_5_12_17_years_old,
    age_18_25 = rc_adult_age_category_recode_1_18_25_years_old,
    age_26_34 = rc_adult_age_category_recode_2_26_34_years_old,
    age_35_49 = rc_adult_age_category_recode_3_35_49_years_old,
    age_50_plus = rc_adult_age_category_recode_4_50_or_older,

    medicaid_coverage = covered_by_medicaid_chip_1_yes,

    less_than_high_school = rc_education_categories_1_less_high_school,
    high_school = rc_education_categories_2_high_school_grad,
    some_college = rc_education_categories_3_some_coll_assoc_dg,
    college_grad = rc_education_categories_4_college_graduate,

    emp_full_time = employment_status_imputation_revised_1_employed_full_time,
    emp_part_time = employment_status_imputation_revised_2_employed_part_time,
    emp_unemployed = employment_status_imputation_revised_3_unemployed,

    pain_relievers_misuse = rc_pain_relievers_past_year_misuse_1_misused_within_the_past_year,

    race_nh_white = rc_race_recode_4_levels_1_white_not_hispanic,
    race_nh_black = rc_race_recode_4_levels_2_black_not_hispanic,
    race_hisp = rc_race_recode_4_levels_4_hispanic,
    race_other = rc_race_recode_4_levels_3_other_or_multiple_not_hispanic,

    poverty = rc_poverty_level_new_inc_percent_of_us_census_poverty_threshold_1_living_in_poverty,
    poverty_2x = rc_poverty_level_new_inc_percent_of_us_census_poverty_threshold_2_income_up_to_2x_fed_pov_thresh,
    poverty_2x_plus = rc_poverty_level_new_inc_percent_of_us_census_poverty_threshold_3_income_more_than_2x_fed_pov_thresh
  ) |> 

  select(
    state,
    year,
    pain_relievers_misuse,
    age_12_17, age_18_25, age_26_34, age_35_49, age_50_plus,
    medicaid_coverage,
    less_than_high_school, high_school, some_college, college_grad,
    emp_full_time, emp_part_time, emp_unemployed,
    race_nh_white, race_nh_black, race_hisp, race_other,
    poverty, poverty_2x, poverty_2x_plus
  )
```

## Exploratory Data Analysis

```{r}
#| label: data-exploration
#| include: false

# Count rows and columns
nrows <- df_renamed |> nrow()
ncols <- df_renamed |> length()

# Get missingness
perc_complete <- round(df_renamed |> n_complete() / nrows / ncols * 100, 2)
```

The data set contains `r nrows` state-year rows and `r ncols` features, and is `r perc_complete` percent complete. The data are all numeric, and the target variable is `pain_relievers_misuse`, which represents the percentage of individuals who misused pain relievers in the past year. The other variables represent various demographic and socioeconomic characteristics of people surveyed in the state and year. See @tbl-summary-statistics for descriptive characteristics of the data.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for all variables in the dataset"
#| echo: true

tbl_summ_stats <- df_renamed |> 
  skim_without_charts() |> 
  janitor::clean_names() |> 
  as_tibble() |>
  rename(Variable = skim_variable, 
         "Completeness rate" = complete_rate, 
         "Mean" = numeric_mean, 
         "SD" = numeric_sd, 
         "P50" = numeric_p50, 
         "Min" = numeric_p0, 
         "P25" = numeric_p25, 
         "P75" = numeric_p75, 
         "Max" = numeric_p100) |>
    select(Variable, "Completeness rate", Mean, Min, P25, P50, P75, Max) |>
    mutate(across(where(is.numeric), ~ signif(., 2))) |> 
  knitr::kable()

tbl_summ_stats
```

```{r}
#| label: fig-correlation-matrix
#| fig-cap: "Correlation matrix of all non-ID variables in the dataset. Correlations < 0.05 are suppressed."
#| fig.align: "center"
#| echo: true
#| warning: false

# Get correlation matrix
df_renamed_cor <- df_renamed |>
  select(-state, -year) |>
  cor(use = "pairwise.complete.obs")

# Suppress small correlations by setting them to NA so corrplot leaves them blank
df_renamed_cor[abs(df_renamed_cor) < 0.05] <- NA

corrplot(df_renamed_cor, 
         method = "circle", 
         type = "upper", 
         tl.cex = 0.4, 
         insig = "blank", 
         na.label = " ", 
         addCoef.col = "black", 
         number.cex = 0.3,
         tl.srt = 45)

most_corr_w_target <- df_renamed_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "pain_relievers_misuse") |>
    transmute(
        Variable,
        Correlation = pain_relievers_misuse,
        Abs_Correlation = abs(pain_relievers_misuse)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.05) |> 
    pull(Variable)
```

The correlation between all variables are shown in @fig-correlation-matrix. The following predictors are most correlated with the target variable: `r paste(most_corr_w_target, collapse = ", ")`, which may be predictors of pain reliever misuse.

Lastly, we also looked at the distribution of each variable to check for skewness and outliers.

```{r}
#| label: fig-distributions
#| fig-cap: "Distribution of numeric variables in the  dataset."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Keep only continuous
continuous_vars <- df_renamed |>
    select(where(is.numeric)) |>
    names()

df_renamed |>
    select(all_of(continuous_vars)) |>
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    facet_wrap(~ variable, scales = "free", ncol = 4) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    labs(x = NULL, y = "Count")
```

Most variables are approximately normally distributed, except for many of the race indicator variables. The proportion of individuals who are non-white--especially those who are an "other" race besides hispanic, non-hispanic white, or non-hispanic black--are skewed right, mostly close to 0 across states and years, with a few higher outliers. This is expected given the racial demographics of the US. The proportion of individuals who are aged 26-34 also has a few high outliers.

## Results of Data Preparation (Alex)

### Transformations

The optimal lambda values for Box-Cox transformation provided insight into which variables would benefit from various types of transformation:

-   `age_50_plus`, `some_college`, `poverty_2x_plus`, and `race_nh_white` had $\lambda \approx 2$, which indicates squared transformation,

-   `emp_unemployed` and `medicaid_coverage` had $\lambda \approx 0.5$, which indicates square root transformation,

-   `race_nh_black`, `race_hisp`, and `poverty`) had $\lambda \approx 0$, which indicates natural log transformation,

-   `less_than_high_school`, `college_grad`, and `race_other` had $\lambda \approx -0.5$, which indicates reciprocal square root transformation,

-   `age_18_25` had $\lambda \approx -1$, which indicates inverse transformation, and

-   `age_26_34` and `age_35_49` had $\lambda = -2$, which indicates inverse squared transformation

Several variables (e.g., `age_12_17`, `poverty_2x`, `high_school`, `emp_full_time`) had $\lambda \approx 1$, which indicates that their distributions are approximately normal, so no (or very little) transformation is needed.

```{r}
#| label: calc-box-cox-lambda

# Numeric predictors
predictors_untransformed <- select(df_renamed, age_12_17:poverty_2x_plus)

# Determine optimal lambda values for Box-Cox transformation
lambdas <- purrr::map_dbl(predictors_untransformed, ~ {
  caret::BoxCoxTrans(.x)$lambda
})

# Convert results vector to dataframe and format
optimal_lambdas <- as.data.frame(lambdas)
optimal_lambdas$variable <- rownames(optimal_lambdas)
rownames(optimal_lambdas) <- NULL
optimal_lambdas |>
  arrange(desc(lambdas)) |>
  select(variable, 'lambda' = 'lambdas') |>
  knitr::kable()
```

These lambda values were then used to apply Box-Cox transformation to the numeric predictor variables.

```{r}
#| label: apply-box-cox-transformation

# Box-Cox transformation of numeric variables using optimal lambda values (previous block)
lambdas_vec <- optimal_lambdas$lambdas
predictors_transformed <- purrr::map2_dfc(
  .x = predictors_untransformed,
  .y = lambdas_vec,
  .f = ~forecast::BoxCox(.x, .y)
)

# Restore the categorical and target variables and create new dataframe
df_prepared <- bind_cols(df_renamed[, 1:3], predictors_transformed)

# Removing the year and state variables since they are not needed for modeling and may cause data leakage if included as predictors
df_prepared <- df_prepared |> select(-state, -year)
```

### Data split

Finally, we split the data 70:30 in preparation for model development.

```{r}
#| label: split-data

set.seed(144)
train_indices <- createDataPartition(df_prepared$pain_relievers_misuse, p = 0.7, list = FALSE)
oud_train_split <- df_prepared[train_indices, ]
oud_test_split <- df_prepared[-train_indices, ]
```

## Results of Assumption checks (Nana)

## Results of Building models (Richie)

Before building the models, we define the 10-fold Cross-Validation logic that we will use for model selection and evaluation. 

```{r}
# Define 10-fold Cross-Validation logic
# Ref: AMAR Ch 7.3 (Predictive Ability)
train_control <- trainControl(method = "cv", number = 10)

# List to contain modeling objects and cross-validation results
model_results_list <- list()
```

Our First model is the Full Multiple Linear Regression model, which includes all predictors. This model serves as a baseline for comparison with more parsimonious models.

```{r}
#| label: model-1-full-mlr

# 1. Full Multiple Linear Regression
model_full <- train(
  pain_relievers_misuse ~ ., 
  data = oud_train_split, 
  method = "lm", 
  trControl = train_control
)

model_full
model_results_list$Full_MLR <- model_full
```

Secondly, we will train our Stepwise Selection model, which iteratively adds or removes predictors based on AIC to find a more parsimonious model.

```{r}
#| label: model-2-stepwise-aic

# 2. Stepwise Selection based on AIC
model_step <- train(
  pain_relievers_misuse ~ ., 
  data = oud_train_split, 
  method = "lmStepAIC", 
  trControl = train_control,
  trace = FALSE
)

model_step
model_results_list$Stepwise_AIC <- model_step
```

Next, we train our LASSO regression model, which applies a penalty to shrink less important coefficients to zero, effectively performing variable selection. We impliment a grid search to find the optimal value of the regularization parameter $\lambda$ that minimizes the cross-validated error.

```{r}
#| label: model-3-lasso-tuning

# 3. LASSO Regression with Hyperparameter Tuning
lasso_grid <- expand.grid(fraction = seq(0.1, 1, length = 10))

model_lasso <- train(
  pain_relievers_misuse ~ ., 
  data = oud_train_split, 
  method = "lasso", 
  trControl = train_control,
  tuneGrid = lasso_grid
)

model_lasso
model_results_list$LASSO <- model_lasso
```

Lastly, we will compare the performance of these three models based on their cross-validated RMSE and select the best performing model for final evaluation on the test set.

```{r}
#| label: model-validation-comparison

# Collecting CV results for comparison
cv_summary <- resamples(list(
  Full = model_results_list$Full_MLR,
  Stepwise = model_results_list$Stepwise_AIC,
  LASSO = model_results_list$LASSO
))

summary(cv_summary)
```

## Results of model selection (Said)

# Discussion and Conclusions (Naomi)

*Conclude your findings, limitations, and suggest areas for future work.*

# References
