---
title: "Predictors of Opioid Misuse Disorder in the US"
subtitle: "Final Project"
format: pdf
bibliography: references.bib
csl: apa.csl
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong
  - name: Said Naqwe
date: April 26, 2026
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Nana Frimpong, Said Naqwe}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: load-packages
#| include: false
# Load necessary libraries
library(tidyverse)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
library(pROC)
library(fastDummies)
library(tidytable)
library(stringr)
```

# Abstract

::: callout-note
**Naomi to complete.** Use 250 words or less.
:::

Background: Opioid addiction is a public health crisis that has affected countless lives.

Research Question: What factors are important for predicting prevalence rate of opioid use disorder (OUD)?

Methods: We downloaded data on the prevalence of pain reliever misuse in each state from the National Survey on Drug Use and Health from the Substance and Mental Health Services Administration for all available years (2016–2019 and 2022)...\[TBD\]

Results: \[TBD\]

Conclusion: \[TBD\]

# Key words

Opioid Use Disorder, Regression Analysis, Public Health, Substance Use Disorder

# Introduction

Opioid addiction is a public health crisis that has affected countless lives. The United States is experiencing a public health crisis of almost unprecedented scale: an epidemic of opioid use disorder (OUD) and related overdose deaths, the magnitude of which has increased exponentially in the past decades [@Mancher2019]. In just one year, the age-adjusted rate of drug overdose deaths increased 31% (from 21.6 in 2019 per 100k population to 28.3 in 2020) [@Hedegaard2021]. In this project, we investigate what factors are important for predicting the prevalence rate of OUD across US states. 

# Literature review

::: callout-note
**Alex to complete.** Discuss how other researchers have addressed similar problems, what their achievements are, and what the advantage and drawbacks of each reviewed approach are. Explain how your investigation is similar or different to the state-of-the- art. Please cite the relevant papers where appropriate.
:::

Example citation [@Volkow2019].

# Methodology

::: callout-note
Discuss the key aspects of your problem, data set and regression model(s). Given that you are working on real-world data, explain at a high-level your exploratory data analysis, how you prepared the data for regression modeling, your process for building regression models, and your model selection.
:::

We use data on the state-level prevalence of pain reliever misuse and demographic characteristics from the National Survey on Drug Use and Health (NSDUH) from the Substance and Mental Health Services Administration for 2018-2024 [@NSDUH2018-2019].

## Exploratory Data Analysis

We performed exploratory data analysis to understand the distribution of our outcome variable (prevalence of pain reliever misuse) and the relationships between our predictors and the outcome. We visualized the correlations between predictors and the outcome variable, as well as the distribution of the numeric variables.

## Data Preparation methods (Alex)

## Assumptions (Nana)

## Build models methods (Richie)

## Select models methods (Said)

# Experimentation and Results

::: callout-note
Describe the specifics of what you did (data exploration, data preparation, model building, model selection, model evaluation, etc.), and what you found out (statistical analyses, interpretation and discussion of the results, etc.).
:::

```{r}
#| label: load-join-data

# Read in all CSVs from the data directory and store them in a list
filepaths <- list.files(path = "data", pattern = "*.csv", full.names = TRUE)

# Helper to extract the "YY-YY" year substring from filenames
extract_year_tag <- function(path) {
  fname <- basename(path)
  tag <- stringr::str_extract(fname, "\\d{2}-\\d{2}")
  if (is.na(tag)) tag <- gsub("\\.csv$", "", fname)
  tag
}

# Read each file and add a 'year' column based on the filename (e.g. "18-19")
data_list <- lapply(filepaths, function(fp) {
  df <- readr::read_csv(fp, show_col_types = FALSE) |> 
    janitor::clean_names()
  
  # Get name of first column (variable of interest)
  first_col <- names(df)[1]

  # Transform each df.
  df <- df |> 
    # Filter out redundant ("Overall") rows
    filter(
      state_us_abbreviation != "Overall") |>

    # Calculate precise percentages
    group_by(state_us_abbreviation) |>
    mutate(
      denominator = weighted_count[column_percent == 1]
    ) |>
    ungroup() |> 
    mutate(
      percent = weighted_count / denominator
    ) |> 

    # Get subset of columns
    select(1, state_us_abbreviation, percent) |> 

    # Pivot first column wider
    mutate(
      !!first_col := paste0(first_col, "_", as.character(.data[[first_col]])),
    ) |>
    pivot_wider(names_from = 1, values_from = percent) |> 

    # Drop columns with all the same values
    select(where(~ n_distinct(.) > 1)) |>

    # Add year column
    mutate(
      year = extract_year_tag(fp)
    ) |>

    # Clean names
    janitor::clean_names()

  df
})

# Get list of data years
years <- sapply(filepaths, extract_year_tag) |> unique()

# Iterate across years, join data frames in data_list by state and year within each year.
yearly_joined_data <- lapply(years, function(yr) {
  year_dfs <- data_list |>
    purrr::keep(~ dplyr::n_distinct(.x$year) == 1 && .x$year[[1]] == yr)

  if (length(year_dfs) == 0) return(NULL)
  if (length(year_dfs) == 1) return(year_dfs[[1]])

  Reduce(function(x, y) full_join(x, y, by = c("state_us_abbreviation", "year")), year_dfs)
})

# Append yearly_joined_data across all years into a single data frame
# Filter out NULL values first to avoid bind_rows errors
df <- yearly_joined_data |> 
  purrr::discard(is.null) |>
  bind_rows()
```

```{r}
#| label: rename-columns

df_renamed <- df |> 
  rename(
    state = state_us_abbreviation,

    age_12_17 = rc_adult_age_category_recode_5_12_17_years_old,
    age_18_25 = rc_adult_age_category_recode_1_18_25_years_old,
    age_26_34 = rc_adult_age_category_recode_2_26_34_years_old,
    age_35_49 = rc_adult_age_category_recode_3_35_49_years_old,
    age_50_plus = rc_adult_age_category_recode_4_50_or_older,

    medicaid_coverage = covered_by_medicaid_chip_1_yes,

    less_than_high_school = rc_education_categories_1_less_high_school,
    high_school = rc_education_categories_2_high_school_grad,
    some_college = rc_education_categories_3_some_coll_assoc_dg,
    college_grad = rc_education_categories_4_college_graduate,

    emp_full_time = employment_status_imputation_revised_1_employed_full_time,
    emp_part_time = employment_status_imputation_revised_2_employed_part_time,
    emp_unemployed = employment_status_imputation_revised_3_unemployed,

    pain_relievers_misuse = rc_pain_relievers_past_year_misuse_1_misused_within_the_past_year,

    race_nh_white = rc_race_recode_4_levels_1_white_not_hispanic,
    race_nh_black = rc_race_recode_4_levels_2_black_not_hispanic,
    race_hisp = rc_race_recode_4_levels_4_hispanic,
    race_other = rc_race_recode_4_levels_3_other_or_multiple_not_hispanic,

    poverty = rc_poverty_level_new_inc_percent_of_us_census_poverty_threshold_1_living_in_poverty,
    poverty_2x = rc_poverty_level_new_inc_percent_of_us_census_poverty_threshold_2_income_up_to_2x_fed_pov_thresh,
    poverty_2x_plus = rc_poverty_level_new_inc_percent_of_us_census_poverty_threshold_3_income_more_than_2x_fed_pov_thresh
  ) |> 

  select(
    state,
    year,
    pain_relievers_misuse,
    age_12_17, age_18_25, age_26_34, age_35_49, age_50_plus,
    medicaid_coverage,
    less_than_high_school, high_school, some_college, college_grad,
    emp_full_time, emp_part_time, emp_unemployed,
    race_nh_white, race_nh_black, race_hisp, race_other,
    poverty, poverty_2x, poverty_2x_plus
  )
```

## Exploratory Data Analysis (Naomi)

```{r}
#| label: data-exploration
#| include: false

# Count rows and columns
nrows <- df_renamed |> nrow()
ncols <- df_renamed |> length()

# Get missingness
perc_complete <- round(df_renamed |> n_complete() / nrows / ncols * 100, 2)
```

The data set contains `r nrows` state-year rows and `r ncols` features, and is `r perc_complete` percent complete. The data are all numeric, and the target variable is `pain_relievers_misuse`, which represents the percentage of individuals who misused pain relievers in the past year. The other variables represent various demographic and socioeconomic characteristics of people surveyed in the state and year. See @tbl-summary-statistics for descriptive characteristics of the data.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for all variables in the dataset"
#| echo: true

tbl_summ_stats <- df_renamed |> skim_without_charts() |> janitor::clean_names() |> as_tibble() |>
    rename(Variable = skim_variable, "Completeness rate" = complete_rate, "Mean" = numeric_mean, "SD" = numeric_sd, "P50" = numeric_p50, "Min" = numeric_p0, "P25" = numeric_p25, "P75" = numeric_p75, "Max" = numeric_p100) |>
    select(Variable, "Completeness rate", Mean, Min, P25, P50, P75, Max) |>
    mutate(across(where(is.numeric), ~ signif(., 2))) |> knitr::kable()

tbl_summ_stats
```

```{r}
#| label: fig-correlation-matrix
#| fig-cap: "Correlation matrix of all non-ID variables in the dataset. Correlations < 0.05 are suppressed."
#| fig.align: "center"
#| echo: true
#| warning: false

# Get correlation matrix
df_renamed_cor <- df_renamed |>
  select(-state, -year) |>
  cor(use = "pairwise.complete.obs")

# Suppress small correlations by setting them to NA so corrplot leaves them blank
df_renamed_cor[abs(df_renamed_cor) < 0.05] <- NA

corrplot(df_renamed_cor, method = "circle", type = "upper", tl.cex = 0.4, insig = "blank", na.label = " ", addCoef.col = "black", number.cex = 0.3, tl.srt = 45)

most_corr_w_target <- df_renamed_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "pain_relievers_misuse") |>
    transmute(
        Variable,
        Correlation = pain_relievers_misuse,
        Abs_Correlation = abs(pain_relievers_misuse)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.05) |> 
    pull(Variable)
```

The correlation between all variables are shown in @fig-correlation-matrix. The following predictors are most correlated with the target variable: `r paste(most_corr_w_target, collapse = ", ")`, which may be predictors of pain reliever misuse.

Lastly, we also looked at the distribution of each variable to check for skewness and outliers. 

```{r}
#| label: fig-distributions
#| fig-cap: "Distribution of numeric variables in the  dataset."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Keep only continuous
continuous_vars <- df_renamed |>
    select(where(is.numeric)) |>
    names()

df_renamed |>
    select(all_of(continuous_vars)) |>
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    facet_wrap(~ variable, scales = "free", ncol = 4) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    labs(x = NULL, y = "Count")

```

Most variables are approximately normally distributed, except for many of the race indicator variables. The proportion of individuals who are non-white--especially those who are an "other" race besides hispanic, non-hispanic white, or non-hispanic black--are skewed right, mostly close to 0 across states and years, with a few higher outliers. This is expected given the racial demographics of the US. The proportion of individuals who are aged 26-34 also has a few high outliers.

## Results of Data Preparation (Alex)

## Results of Assumption checks (Nana)

## Results of Building models (Richie)

## Results of model selection (Said)

# Discussion and Conclusions (Naomi)

*Conclude your findings, limitations, and suggest areas for future work.*

# References