---
title: "Classification Metrics"
subtitle: "Homework #2"
format: pdf
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong 
  - name: Said Naqwe
date: March 22, 2026fi
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Said Naqwe, and Nana Frimpong}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: Load necessary libraries
#| include: TRUE
#| warning: FALSE
library(tidyverse)
library(skimr)
library(corrplot)
library(ggplot2)
library(GGally)
library(janitor)
```

# Homework Solutions

1.  *Download the classification output data set (attached in Blackboard to the assignment).*

    ```{r}
    #| label: Get and summarize data
    classifications_output_data <- read.csv("classification-output-data (2).csv")

    classifications_output_data |> skim()
    ```

2.  *The data set has three key columns we will use:*

-   *`class`: the actual class for the observation*

-   *`scored.class`: the predicted class for the observation (based on a threshold of 0.5)*

-   *`scored.probability`: the predicted probability of success for the observation*

    *Use the table() function to get the raw confusion matrix for this scored dataset. Make sure you understand the output. In particular, do the rows represent the actual or predicted class? The columns?*

    ```{r}
    #| label: Confusion matrix
    confusion_matrix <- table(
    classifications_output_data$class,
    classifications_output_data$scored.class
    )
    confusion_matrix
    ```

    ```{r}
    #| label: True negatives and false negatives
    #| include: TRUE
    tn <- confusion_matrix[1, 1]
    fn <- confusion_matrix[2, 1]
    
    # print
    print(paste("True Negatives:", tn))
    print(paste("False Negatives:", fn))
    ```

    The rows represent the actual class (`class`) and the columns represent the predicted class (`scored.class`). For example, there are `r tn` observations where the actual class is 0 and the predicted class is also 0 (true negatives), and there are `r fn` observations where the actual class is 1 but the predicted class is 0 (false negatives).

3.  *Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the accuracy of the predictions.*

    $$
    𝐴𝑐𝑐𝑢𝑟𝑎𝑐𝑦 = \frac{𝑇𝑃 + 𝑇𝑁}{𝑇𝑃 + 𝐹𝑃 + 𝑇𝑁 + 𝐹𝑁}
    $$

    ```{r}
    #| label: Accuracy function

    calculate_accuracy <- function(data, actual_col, predicted_col) {
    confusion_matrix <- table(data[[actual_col]], data[[predicted_col]])
    tp <- confusion_matrix[2, 2]
    tn <- confusion_matrix[1, 1]
    fp <- confusion_matrix[1, 2]
    fn <- confusion_matrix[2, 1]
    accuracy <- (tp + tn) / (tp + fp + tn + fn)
    return(accuracy)
    }

    # Test function
    accuracy <- calculate_accuracy(classifications_output_data, "class", "scored.class")
    print(paste("Accuracy:", accuracy))
    ```

4.  *Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the classification error rate of the predictions.*

    $$
    𝐶𝑙𝑎𝑠𝑠𝑖𝑓𝑖𝑐𝑎𝑡𝑖𝑜𝑛 𝐸𝑟𝑟𝑜𝑟 𝑅𝑎𝑡𝑒 = \frac{𝐹𝑃 + 𝐹𝑁}{𝑇𝑃 + 𝐹𝑃 + 𝑇𝑁 + 𝐹𝑁}
    $$

    *Verify that you get an accuracy and an error rate that sums to one.*

    ```{r}
    calculate_error_rate <- function(data, actual_col, predicted_col) {
    confusion_matrix <- table(data[[actual_col]], data[[predicted_col]])
    tp <- confusion_matrix[2, 2]
    tn <- confusion_matrix[1, 1]
    fp <- confusion_matrix[1, 2]
    fn <- confusion_matrix[2, 1]
    error_rate <- (fp + fn) / (tp + fp + tn + fn)
    return(error_rate)
    }

    # Test function
    error_rate <- calculate_error_rate(classifications_output_data, "class", "scored.class")
    print(paste("Error Rate:", error_rate))

    # Verify that accuracy and error rate sum to 1
    print(paste("Sum of accuracy + error rate:", accuracy + error_rate))
    ```

5.  *Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the precision of the predictions.*

    $$
    𝑃𝑟𝑒𝑐𝑖𝑠𝑖𝑜𝑛 = \frac{𝑇𝑃}{𝑇𝑃 + 𝐹𝑃}
    $$

    ```{r}

# Precision function

calculate_precision <- function(data, actual_col, predicted_col) {
  confusion_matrix <- table(data[[actual_col]], data[[predicted_col]])
  
  tp <- confusion_matrix[2, 2]
  tn <- confusion_matrix[1, 1]
  fp <- confusion_matrix[1, 2]
  fn <- confusion_matrix[2, 1]
  
  precision <- tp / (tp + fp)
  
  return(precision)
}

# Test function
precision <- calculate_precision(classifications_output_data, "class", "scored.class")
print(paste("Precision:", precision))
    ```

6.  *Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the sensitivity of the predictions. Sensitivity is also known as recall.*

    $$
    𝑆𝑒𝑛𝑠𝑖𝑡𝑖𝑣𝑖𝑡𝑦 = \frac{𝑇𝑃}{𝑇𝑃 + 𝐹𝑁}
    $$

    ```{r}

# Sensitivity function

calculate_sensitivity <- function(data, actual_col, predicted_col) {
  confusion_matrix <- table(data[[actual_col]], data[[predicted_col]])
  
  tp <- confusion_matrix[2, 2]
  tn <- confusion_matrix[1, 1]
  fp <- confusion_matrix[1, 2]
  fn <- confusion_matrix[2, 1]
  
  sensitivity <- tp / (tp + fn)
  
  return(sensitivity)
}

# Test function
sensitivity <- calculate_sensitivity(classifications_output_data, "class", "scored.class")
print(paste("Sensitivity:", sensitivity))
    ```

7.  *Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the specificity of the predictions.*

    $$
    𝑆𝑝𝑒𝑐𝑖𝑓𝑖𝑐𝑖𝑡𝑦 = \frac{𝑇𝑁}{𝑇𝑁 + 𝐹𝑃}
    $$

    ```{r}
    calculate_specificity <- function(data, actual_col, predicted_col) {
    confusion_matrix <- table(data[[actual_col]], data[[predicted_col]])
    tp <- confusion_matrix[2, 2]
    tn <- confusion_matrix[1, 1]
    fp <- confusion_matrix[1, 2]
    fn <- confusion_matrix[2, 1]
    specificity <- tn / (tn + fp)
    return(specificity)
    }

    # Test function
    specificity <- calculate_specificity(classifications_output_data, "class", "scored.class")
    print(paste("Specificity:", specificity))
    ```

8.  *Write a function that takes the data set as a dataframe, with actual and predicted classifications identified, and returns the F1 score of the predictions.*

    $$
    𝐹1 𝑆𝑐𝑜𝑟𝑒 = \frac{2 × 𝑃𝑟𝑒𝑐𝑖𝑠𝑖𝑜𝑛 × 𝑆𝑒𝑛𝑠𝑖𝑡𝑖𝑣𝑖𝑡𝑦}{𝑃𝑟𝑒𝑐𝑖𝑠𝑖𝑜𝑛 + 𝑆𝑒𝑛𝑠𝑖𝑡𝑖𝑣𝑖𝑡𝑦}
    $$

    ```{r}

# F1 score function

calculate_f1 <- function(data, actual_col, predicted_col) {
  
  precision <- calculate_precision(data, actual_col, predicted_col)
  sensitivity <- calculate_sensitivity(data, actual_col, predicted_col)
  
  f1_score <- (2 * precision * sensitivity) / (precision + sensitivity)
  
  return(f1_score)
}

# Test function
f1_score <- calculate_f1(classifications_output_data, "class", "scored.class")
print(paste("F1 Score:", f1_score))
    ```

9.  *Before we move on, let’s consider a question that was asked: What are the bounds on the F1 score? Show that the F1 score will always be between 0 and 1. (Hint: If 0 \< 𝑎 \< 1 and 0 \< 𝑏 \< 1 then 𝑎𝑏 \< 𝑎.)*

    ```{r}

# F1 bounds demonstration

# Example values between 0 and 1
precision_test <- 0.7
sensitivity_test <- 0.8

f1_test <- (2 * precision_test * sensitivity_test) / 
           (precision_test + sensitivity_test)

print(paste("Example F1 Score:", f1_test))
    ```

10. *Write a function that generates an ROC curve from a data set with a true classification column (class in our example) and a probability column (scored.probability in our example). Your function should return a list that includes the plot of the ROC curve and a vector that contains the calculated area under the curve (AUC). Note that I recommend using a sequence of thresholds ranging from 0 to 1 at 0.01 intervals.*

    ```{r}

    ```

11. *Use your created R functions and the provided classification output data set to produce all of the classification metrics discussed above.*

    ```{r}

    ```

12. *Investigate the caret package. In particular, consider the functions confusionMatrix, sensitivity, and specificity. Apply the functions to the data set. How do the results compare with your own functions?*

    ```{r}
    #| label: Question 12 - caret package

    # Load caret package
    library(caret)

    # Create confusion matrix using caret
    cm_caret <- confusionMatrix(as.factor(classifications_output_data$scored.class), 
                                as.factor(classifications_output_data$class))

    print(cm_caret)

    # Extract metrics from caret
    caret_accuracy <- cm_caret$overall['Accuracy']
    caret_sensitivity <- cm_caret$byClass['Sensitivity']
    caret_specificity <- cm_caret$byClass['Specificity']

    print(paste("Caret Accuracy:", caret_accuracy))
    print(paste("Caret Sensitivity:", caret_sensitivity))
    print(paste("Caret Specificity:", caret_specificity))

    # Compare with our functions
    print("Comparison:")
    print(paste("Accuracy - Ours:", accuracy, "Caret:", caret_accuracy))
    print(paste("Sensitivity - Ours:", sensitivity, "Caret:", caret_sensitivity))
    print(paste("Specificity - Ours:", specificity, "Caret:", caret_specificity))
    ```

13. *Investigate the pROC package. Use it to generate an ROC curve for the data set. How do the results compare with your own functions?*

    ```{r}
    #| label: Question 13 - pROC package
    
    # Load pROC package
    library(pROC)

    # Generate ROC curve using pROC
    roc_obj <- roc(classifications_output_data$class, classifications_output_data$scored.probability)

    # Plot the ROC curve
    plot(roc_obj, main = "ROC Curve using pROC")

    # Get AUC
    auc_proc <- auc(roc_obj)
    print(paste("AUC from pROC:", auc_proc))

    # Compare with our custom function (assuming roc_result$auc is available from section 10)
    print("Comparison:")
    print(paste("Custom AUC:", roc_result$auc, "pROC AUC:", auc_proc))
    ```
