---
title: "LOGISTIC REGRESSION"
subtitle: "Homework #3"
format: pdf
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong
  - name: Said Naqwe
date: April 12, 2026
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Nana Frimpong, Said Naqwe}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: load-packages
#| include: false

# Check our current directory and change if necessary
current_dir <- getwd()

# Use grepl to see if './DATA621/' is in the current path
if (grepl("DATA621", current_dir)) {
  # Set the new directory
  setwd("./homework-3/")
  message("Changed directory to: ", getwd())
} else {
  message("Condition not met. Staying in: ", current_dir)
}

# Load necessary libraries
library(tidyverse)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
```

# Introduction

We build a binary logistic regression model on a data set containing information on crime for various neighborhoods of a major city to predict whether the neighborhood will be at risk for high crime levels. Each record in our data has a response variable indicating whether or not the crime rate is above the median crime rate or not. Our R statistical programming code is included in Appendix A.

# Data Exploration

::: callout-note
Data exploration section completed.
:::

```{r}
#| label: data-exploration
#| include: false

# Load the training and evaluation datasets
crime_train <- read_csv("./data/crime-training-data_modified.csv")
crime_eval <- read_csv("./data/crime-evaluation-data_modified.csv")

# Count rows and columns
nrows <- crime_train |> nrow()
ncols <- crime_train |> length()

# Get missingness
perc_complete <- round(crime_train |> n_complete() / nrows / ncols * 100, 2)
```

The crime training data set contains `r nrows` rows and `r ncols` features, and is `r perc_complete` percent complete. The data are all numeric, and the target variable is `target`, which represents whether the crime rate is above the median crime rate (1) or not (0). The other variables represent various information on crime for neighborhoods of a major city. See Appendix B for a full list of variable names and definitions. @tbl-summary-statistics presents descriptive characteristics of the training data.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for all variables in the training dataset"
#| echo: false

tbl_summ_stats <- crime_train |> skim_without_charts() |> janitor::clean_names() |> as_tibble() |>
    rename(Variable = skim_variable, "Completeness rate" = complete_rate, "Mean" = numeric_mean, "SD" = numeric_sd, "P50" = numeric_p50, "Min" = numeric_p0, "P25" = numeric_p25, "P75" = numeric_p75, "Max" = numeric_p100) |>
    select(Variable, "Completeness rate", Mean, Min, P25, P50, P75, Max) |>
    mutate(across(where(is.numeric), ~ signif(., 2))) |> knitr::kable()

tbl_summ_stats
```

```{r}
#| label: fig-correlation-matrix
#| fig-cap: "Correlation matrix of all variables in the training dataset"
#| fig.align: "center"
#| echo: false
#| warning: false

# Get correlation matrix
crime_train_cor <- crime_train |>
    cor(use = "pairwise.complete.obs")
    
corrplot(crime_train_cor, method = "circle", type = "upper", tl.cex = .6, sig.level = 0.05, insig = "blank", addCoef.col = "black", number.cex = .5)

most_corr_w_target <- crime_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable == "target") |>
    mutate(across(where(is.numeric), ~ signif(., 2)))
```

The correlation between all variables are shown @fig-correlation-matrix. The most correlated predictor variables with `target` are `nox` (nitrogen oxides concentration, R = `r most_corr_w_target$nox`), `age` (proportion of owner-occupied units built prior to 1940, R = `r most_corr_w_target$age`), `rad` (index of accessibility to radial highways, R = `r most_corr_w_target$rad`), `dis` (weighted mean of distances to five Boston employment centers, R = `r most_corr_w_target$dis`), `tax` (full-value property-tax rate per \$10,000, R = `r most_corr_w_target$tax`), and `indus` (proportion of non-retail business acres per suburb, R = `r most_corr_w_target$indus`). These variables have the strongest linear relationships with the target variable and may be good predictors of crime rates.

```{r}
#| label: tbl-most-correlated-predictors
#| tbl-cap: "Multicollinear variables in the training dataset"
#| echo: false

crime_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Predictor 1") |>
    pivot_longer(-`Predictor 1`, names_to = "Predictor 2", values_to = "correlation") |>
    filter(`Predictor 1` != `Predictor 2`) |>
    distinct(across(c(correlation)), .keep_all = TRUE) |>
    arrange(desc(abs(correlation))) |>
    rename(Correlation = correlation) |>
    filter(abs(Correlation) > 0.7) |>
    mutate(across(where(is.numeric), ~ signif(., 2))) |>
    knitr::kable()
```

@tbl-most-correlated-predictors shows predictor pairs with a pairwise correlation coefficient greater than 0.7. High pairwise correlation can create modeling issues because it becomes harder to isolate each predictor's individual effect on the target variable. The strongest multicollinear pair is `rad` (accessibility to highways) and `tax` (property-tax rate), which is plausible because neighborhoods with better highway access may also have higher property taxes. For this pair, and for the other pairs in the table, we should consider dropping one variable or using regularization techniques to mitigate multicollinearity.

```{r}
#| label: fig-box-plots
#| fig-cap: "Box plots for identifying outliers in the training data"
#| fig.align: "center"
#| echo: false
#| warning: false

crime_train |>
    pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") |>
    group_by(Variable) |>
    mutate(Z_Score = scale(Value)[,1]) |>
    ungroup() |>
    ggplot(aes(x = Z_Score, y = Variable)) +
    geom_boxplot() +
    theme_minimal() +
    theme(panel.grid = element_blank(), axis.text.y = element_text(angle = 0)) +
    xlab("Z-Score")
```

@fig-box-plots shows the distribution of z-scores for all variables and highlights potential outliers in the training data. The variable `zn` (proportion of residential land zoned for large lots over 25,000 square feet) has many high outliers, which is plausible because some neighborhoods may have much higher shares of large lots than others; its maximum value is `r max(crime_train$zn)` percent. The variable `rm` (average number of rooms per dwelling) also has several outliers, ranging from `r min(crime_train$rm)` to `r max(crime_train$rm)` rooms. Likewise, `medv` (median value of owner-occupied homes) ranges from \$`r min(crime_train$medv)`k to \$`r max(crime_train$medv)`k, which is not unexpected given variation in housing prices across neighborhoods and unspecified data years. The variables `lstat` (percent lower status population) and `dis` (weighted mean distance to five Boston employment centers) show a few high outliers (`r max(crime_train$lstat)` percent and `r max(crime_train$dis)` units, respectively). The variable `chas` has a high value of 1, which is expected because it is a binary indicator of whether a suburb borders the Charles River. Overall, these outliers appear to reflect real neighborhood variability rather than data-entry errors. If influence diagnostics later reveal problematic leverage points, we can consider adding indicator variables to account for them.

# Data Preparation

::: callout-note
Completed

I don't think we need to compare this dataset to the baseball dataset (Homework 1).
:::

For the crime data used in Homework 3, the first step in data preparation was to check whether any variables contained missing values. This crime data set does not contain missing values in either the training or evaluation files. Therefore, no imputation or missingness flag variables were needed.

Even though missing-value treatment was unnecessary, we still prepared the data for binary logistic regression by creating transformed and derived predictors. Several predictors are right-skewed, especially `zn`, `dis`, `rad`, `tax`, and `lstat`. For these variables, log transformations were created using `log1p()`, which reduces skewness while safely handling zero values. These transformed variables may improve model fit by making relationships with the log-odds of the response more stable.

We also created a small set of engineered variables that may be useful in later models. First, `nox_indus` captures the combined effect of air pollution concentration and industrial land use. Second, `rm_lstat_ratio` measures the balance between average rooms per dwelling and lower-status population, combining a housing-quality measure with a socioeconomic measure. Third, `tax_rad_ratio` captures the relationship between tax burden and highway accessibility. These new variables were created because logistic regression can benefit from transformed and combined predictors when the original variables may not fully capture the structure related to high-crime neighborhoods.

Finally, the cleaned training and evaluation data sets were stored as `crime_clean` and `crime_eval_clean` so the later modeling section can use them directly.

```{r}
#| label: data-preparation-check-missing
#| echo: true

# Check missing values in training and evaluation data
missing_train <- crime_train |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing")

missing_eval <- crime_eval |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing")

missing_train
missing_eval

#| label: data-preparation-transformations
#| echo: true
#| warning: false

# Start with copies of the original datasets
crime_clean <- crime_train
crime_eval_clean <- crime_eval

# Create transformed and engineered predictors
crime_clean <- crime_clean |>
  mutate(
    # Log transforms for right-skewed predictors
    log_zn    = log1p(zn),
    log_dis   = log1p(dis),
    log_rad   = log1p(rad),
    log_tax   = log1p(tax),
    log_lstat = log1p(lstat),

    # Engineered variables
    nox_indus     = nox * indus,
    rm_lstat_ratio = rm / lstat,
    tax_rad_ratio  = tax / rad
  )

crime_eval_clean <- crime_eval_clean |>
  mutate(
    # Log transforms for right-skewed predictors
    log_zn    = log1p(zn),
    log_dis   = log1p(dis),
    log_rad   = log1p(rad),
    log_tax   = log1p(tax),
    log_lstat = log1p(lstat),

    # Engineered variables
    nox_indus     = nox * indus,
    rm_lstat_ratio = rm / lstat,
    tax_rad_ratio  = tax / rad
  )

#| label: data-preparation-verify
#| echo: true

# Verify dimensions after preparation
dim(crime_clean)
dim(crime_eval_clean)

# Verify that no missing values remain after transformations
crime_clean |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") |>
  filter(Missing > 0)

crime_eval_clean |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") |>
  filter(Missing > 0)


### Linearity Assumption Summary

# The results above indicate that the assumption of linearity is satisfied for all independent variables. However, several were flagged as being unlikely to have predictive value in a multivariate linear model.

## Data Splits

# First, we will split our dataset into a training and testing set. We will use the training set to build our models and the testing set to evaluate their performance. The training set will be used to fit the models, while the testing set will be used to assess how well the models generalize to new data.

#| label: split-data
#| eval: false

set.seed(19940211)
train_indices <- sample(nrow(crime_clean), size = 0.8 * nrow(crime_clean))
crime_train_split <- crime_clean[train_indices, ]
crime_test_split <- crime_clean[-train_indices, ]
```

# Check assumptions for binary logistic regression

::: callout-note
To be added
:::

# Build models

::: callout-note
Completed but file won't render. The rendering fails on line 265, but I'm not sure what's wrong. Any ideas?
:::

We compared three models. Our first model included all the variables in the cleaned training dataset. The second model aimed to reduce multicollinearity by omitting variables that had correlation coefficients $|r|>0.7$ (see Data Exploration section). Finally, we generated the third model using backward stepwise selection from a full model that included transformed variables and newly created variables (see Data Preparation section).

## Model 1. All untransformed variables

As shown below, the most significant predictors were `nox` and `rad` ($P<0.001$). In addition, `age` and `ptratio` were significant at the $P<0.01$ level, and `dis` and `tax` were significant at the $P<0.05$ level. The AIC for this model is `r\ AIC(model1)`.

```{r}
#| label: build-model1

model1 <- glm(
  formula = target ~ zn + indus + chas + nox + rm + age + dis + rad + tax +
                     ptratio + lstat + medv,
  family = "binomial",
  data = crime_train_split)

summary(model1)
```

## Model 2. Reduce multicollinearity

In the Data Exploration section, we showed that `rad` and `tax` were strongly correlated (r=0.91), which suggests that only one of these variables is needed in Model 1 (we chose to retain `rad` as it was more highly significant). Similarly, `medv` was shown to be potentially multicollinear with `lstat` (r=0.74) and `rm` (r=0.71), so we removed the latter variables.

In addition, the correlation between `nox` (predictor) and `target` (response) suggests that `nox` is an important predictor, so we removed its potential collinear pairs `age` and `dis`. Although `nox` was moderately correlated with `indus` (r=0.76), we retained `indus` because removal of both `indus` and `tax` (highly correlated with `rad`) would eliminate a potentially informative predictor.

These changes resulted in a model with AIC `r\ AIC(model2)`.

```{r}
#| label: build-model2

model2 <- glm(
  formula = target ~ zn + indus + chas + nox + rad + ptratio + medv,
  family = "binomial",
  data = crime_train_split)

summary(model2)
```

## Model 3. Backward stepwise variable selection

For this model, we performed backward stepwise selection from the set of variables including transformed and newly created variables.

The best model had `nox`, `tax_rad_ratio`, `ptratio`, and `age` as the most significant predictors (P\<0.001). In addition, `log_dis` and `rm_lstat_ratio` were significant at the P\<0.05 level. The AIC of this model was `r\ AIC(model3)`, which was lower than that of Models 1 and 2.

```{r}
#| label: build-model3

# Full model includes all variables
model_full <- glm(
  formula = target ~ log_zn + indus + chas + nox + rm + age + 
                     log_dis + log_rad + log_tax + ptratio + log_lstat + medv +
                     nox_indus + rm_lstat_ratio + tax_rad_ratio,
  family = "binomial",
  data = crime_train_split)

# Select variables
model3 <- step(model_full, direction = "backward")

summary(model3)
```

Of these three models, model 3 has the lowest AIC. In the next section, we will evaluate the performance of these models to more determine which one is best for predicting the `target` variable.

# Select Models

```{r}
#| label: model-comparison-table
#| echo: true
#| results: asis
#| eval: false

# Creating a custom helper function to extract all the required metrics 
# explicitly evaluating on the provided training dataset.
get_model_metrics <- function(model, model_name, train_data) {
  mod_summary <- summary(model)
  
  # Explicitly calculate Mean Squared Error (MSE) using the training data
  predictions <- predict(model, newdata = train_data)
  mse <- mean((train_data$TARGET_WINS - predictions)^2)
  
  # Extract F-statistic and calculate its p-value (inherent to training data)
  f_stat <- mod_summary$fstatistic
  p_val <- if (!is.null(f_stat)) {
    pf(f_stat[1], f_stat[2], f_stat[3], lower.tail = FALSE)
  } else {
    NA
  }
  
  # Return a nicely formatted row for our table
  tibble(
    Model = model_name,
    `Training MSE` = mse,
    `R-Squared` = mod_summary$r.squared,
    `Adj R-Squared` = mod_summary$adj.r.squared,
    `F-Statistic` = f_stat[1],
    `P-Value` = p_val,
    `AIC` = AIC(model),
    `BIC` = BIC(model)
  )
}

# Apply the function to all three models, explicitly passing the training split
model_comparison <- bind_rows(
  get_model_metrics(model_full, "Full Model", crime_train_split),
  get_model_metrics(model_theory, "Theory Model", crime_train_split),
  get_model_metrics(model_stepwise, "Stepwise Model", crime_train_split)
) |>
  # Sort by the lowest Training MSE 
  arrange(`Training MSE`)

# Display the table
knitr::kable(
  model_comparison, 
  digits = 4, 
  caption = "Evaluation of Candidate Models on Training Dataset"
)
```

We favor a slightly more parsimonious model over a complex one with a marginally better R2. A simpler model (like our Stepwise or Theory model) is easier to interpret and less prone to overfitting. We will evaluate the models using Adjusted `R^2` and MSE on the training data, and then use the evaluation dataset to compare their predictive performance.

Looking at the table above, we can see that the Stepwise and Full models have similar MSE and Adjusted `R^2`, but the Stepwise model has fewer predictors making it a bit more parsimonious. Additionally, the Stepwise model is more interpretable and less prone to overfitting than the Full model. Additonally, the theory model has a much higher MSE and lower Adjusted `R^2` than the other two models, which suggests that baseball is more comoplex to predict wins with just the most correlated variables and the two new predictors. Therefore, **we select the Stepwise model as our final model for predicting `TARGET_WINS`**.

Regarding the F-statistic, the highly significant p-value for the stepwise model's F-statistic (119.9) confirms that our selected group of predictors is collectively highly effective at explaining the variance in total team wins.

```{r}
#| label: stepwise-model-summary
#| echo: true
#| eval: false
# Display the summary of the selected Stepwise model to interpret coefficients and significance
summary(model_stepwise)
```

To make inferences from our Stepwise model, we can interpret the coefficients. For example, the coefficient for `TEAM_BATTING_H` is 0.05 which indicates that, all other things being equal, for every additional 20 hits a team records over a season, they're expected to gain about 1 additional win. Similarly, `TEAM_BATTING_BB` (i.e., walks) has a positive highly significant coefficient (0.03) which aligns with baseball theory that teams that walk more tend to win more.

# Results

```{r}
#| label: predictions-on-eval-set
#| echo: true
#| eval: false
# Generate predictions on the evaluation dataset using the selected model
eval_predictions <- predict(model_stepwise, newdata = crime_eval_clean)

# Bind the predictions back to the evaluation dataset for easy viewing
crime_eval_results <- crime_eval_clean |>
  # We round the predictions to the nearest whole number because a team 
  # can't win a fraction of a baseball game
  mutate(PREDICTED_TARGET_WINS = round(eval_predictions, 0)) |>
  # Move the new prediction column to the front (right after INDEX) for easy reading
  relocate(PREDICTED_TARGET_WINS, .after = INDEX)

# Write predictions to a csv
write_csv(crime_eval_results, "crime_evaluation_predictions.csv")

# Display the first 10 predictions in the rendered PDF
knitr::kable(
  head(crime_eval_results |> select(INDEX, PREDICTED_TARGET_WINS), 10),
  caption = "Excerpts of Stepwise Model Win Predictions for Evaluation Set"
)
```

Given our evaluation set and our stepwise model, we are able to create predictions for the number of wins for the team in each year. The first 10 predictions are shown in the table above. We write these predictions to a CSV file for submission.

# Conclusion

In conclusion, our stepwise model performed best in predicting the number of baseball wins based on the training data. Since there were some minor violations of assumptions, these results should be interpreted with caution, which may affect the reliability of our model. Future work could explore data transformations or alternative modeling approaches to address these assumption violations and potentially improve predictive performance.

# Appendix A: R Code

::: callout-note
If we want to include our R statistical programming code in an Appendix, add all code here and set the chunk option `include = false` to hide the code from the main report.
:::

# Appendix B: Data Dictionary

| Variable | Description | Role |
|------------------------|------------------------|------------------------|
| `zn` | Proportion of residential land zoned for large lots (over 25,000 square feet) | Predictor |
| `indus` | Proportion of non-retail business acres per suburb | Predictor |
| `chas` | Dummy variable for whether the suburb borders the Charles River (1) or not (0) | Predictor |
| `nox` | Nitrogen oxides concentration (parts per 10 million) | Predictor |
| `rm` | Average number of rooms per dwelling | Predictor |
| `age` | Proportion of owner-occupied units built prior to 1940 | Predictor |
| `dis` | Weighted mean of distances to five Boston employment centers | Predictor |
| `rad` | Index of accessibility to radial highways | Predictor |
| `tax` | Full-value property-tax rate per \$10,000 | Predictor |
| `ptratio` | Pupil-teacher ratio by town | Predictor |
| `lstat` | Lower status of the population (percent) | Predictor |
| `medv` | Median value of owner-occupied homes in \$1000s | Predictor |
| `target` | Whether the crime rate is above the median crime rate (1) or not (0) | Response |
