---
title: "Wine"
subtitle: "Homework #5"
format: pdf
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong
  - name: Said Naqwe
date: May 16, 2026
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Nana Frimpong, Said Naqwe}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: load-packages
#| include: false

# Load necessary libraries
library(tidyverse)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
library(caret)
library(pROC)
library(fastDummies)
library(tidytable)
library(readr)
library(dplyr)
library(MASS)
library(car)
library(glmnet)
```

# Introduction

In this homework assignment, we explore, analyze, and model a data set containing information on approximately 12,000 commercially available wines. Our objective is to build a count regression model to predict the number of cases of wine that will be sold given certain properties of the wine.

# Data Exploration

```{r}
#| label: data-exploration
#| warning: false

# Load the training and evaluation datasets
wine_train <- read_csv("./data/wine-training-data.csv")
wine_eval <- read_csv("./data/wine-evaluation-data.csv")

# Count rows and columns
nrows <- wine_train |> nrow()
ncols <- wine_train |> length()

# Get missingness
perc_complete <- round(wine_train |> n_complete() / nrows / ncols * 100, 2)
```

The wine training data set contains `r nrows` rows and `r ncols` features, and is `r perc_complete` percent complete. We will need to address the missingness in the data preparation section. All variables are numeric; most are continuous variables describing chemical characteristics of the wines, while others like `STARS` and `LabelAppeal` are discrete. The target variable is `TARGET`, which represents the number of cases purchased.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for all variables in the training dataset"
#| echo: false

tbl_summ_stats <- wine_train |>
    skim_without_charts() |>
    janitor::clean_names() |>
    as_tibble() |>
    rename(
      Variable = skim_variable, 
      "Completeness rate" = complete_rate, 
      "Mean" = numeric_mean, 
      "SD" = numeric_sd, 
      "P50" = numeric_p50, 
      "Min" = numeric_p0, 
      "P25" = numeric_p25, 
      "P75" = numeric_p75, 
      "Max" = numeric_p100) |>
    dplyr::select(Variable, "Completeness rate", Mean, Min, P25, P50, P75, Max) |>
    mutate(
        across(where(is.numeric), ~ signif(., 2)),
        Mean = format(Mean, scientific = FALSE, trim = TRUE)
    ) |>
    knitr::kable()

tbl_summ_stats
```

@tbl-summary-statistics presents descriptive characteristics of the training data.

```{r}
#| label: fig-correlation-matrix
#| fig-cap: "Correlation matrix of all variables in the training dataset. Note: Correlations with absolute value less than 0.05 are suppressed."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Get correlation matrix
wine_train_cor <- wine_train |>
    dplyr::select(where(is.numeric)) |> 
    cor(use = "pairwise.complete.obs")

wine_train_cor_plot <- wine_train_cor
wine_train_cor_plot[abs(wine_train_cor_plot) < 0.05] <- NA

corrplot(wine_train_cor_plot, method = "circle", type = "upper", sig.level = 0.05, insig = "blank", addCoef.col = "black", number.cex = 0.7, tl.cex = 0.7, tl.srt = 45, na.label = " ")

most_corr_w_target_flag <- wine_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "TARGET") |>
    transmute(
        Variable,
        Correlation = TARGET,
        Abs_Correlation = abs(TARGET)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.05) |> 
    pull(Variable)

```

The correlation between all variables are shown @fig-correlation-matrix. The variables most correlated with `TARGET`, the number of cases purchased, are `r paste(most_corr_w_target_flag, collapse = ", ")`. This means that wines with higher ratings (`STARS`), more appealing labels (`LabelAppeal`), lower acidity (`AcidIndex` and `VolatileAcidity`), higher alcohol content (`Alcohol`), and higher sulfur dioxide levels (`TotalSulfurDioxide`) tend to have higher sales. These variables will be important to consider in our modeling process, as they may have strong predictive power for the target variable. 

It is important to note that label appeal and star ratings are likely to be highly correlated with each other, which may lead to multicollinearity issues in a regression model. Similarly, `AcidIndex` and `VolatileAcidity` are likely to be correlated, which may also lead to multicollinearity issues. We will need to consider this when selecting variables for our model.

Lastly, we also looked at the distribution of each variable to check for skewness and outliers. 

```{r}
#| label: fig-distributions
#| fig-cap: "Distribution of continuous variables in the training dataset."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Keep only continuous
continuous_vars <- wine_train |>
    dplyr::select(where(is.numeric), -INDEX) |>
    names()

wine_train |>
    dplyr::select(all_of(continuous_vars)) |>
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    facet_wrap(~ variable, scales = "free", ncol = 4) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    labs(x = NULL, y = "Count")

```

`AcidIndex` is right-skewed, with fewer lines having high acidity. Many other variables appear normally distributed, such as `Alcohol`, `pH`, and `TotalSulfurDioxide`. `STARS` is a discrete variable with values from 1 to 4, and is right-skewed with more wines having lower ratings. `LabelAppeal` is also discrete, with values from -2 to 2, and is normally distributed around 0. There do not appear to be any extreme outliers in the data, although many variables have high frequency of median values, which may indicate that many wines have similar characteristics in these dimensions. Note that `TARGET` has a high frequency of 0 values, which may indicate that many wines have no sales. This may require special consideration in our modeling process.

# Data Preparation

For the wine data used in this assignment, the first step in data preparation was to examine whether any variables contained missing values. Unlike the crime dataset from Homework 3, the wine dataset contains missing values across several predictor variables in both the training and evaluation datasets. Because of this, appropriate missing-value treatment was required before proceeding to modeling.

To address missing data, median imputation was applied to all numeric predictor variables. Median imputation was selected because many of the chemical variables (such as sulfur dioxide levels, residual sugar, and alcohol content) are skewed and may contain outliers. Using the median provides a robust estimate that is less sensitive to extreme values than the mean.

In addition to imputation, missingness indicator variables were created for each predictor. This was done because the assignment specifically notes that missing values themselves may be predictive of the response variable. By creating binary flags indicating whether a value was originally missing, we preserve this potential signal for use in the modeling stage.

Next, several transformations were applied to improve the distribution of the predictors. Variables such as `alcohol`, `residualsugar`, `freesulfurdioxide`, and `totalsulfurdioxide` exhibited right-skewed distributions, so log transformations were created using `log1p()`. These transformations reduce skewness and help stabilize relationships between predictors and the count response variable.

We also created a set of engineered variables to capture relationships between predictors that may not be fully represented by the original variables. The variable `sulfur_ratio` measures the relationship between total and free sulfur dioxide, while `acidity_ratio` compares volatile acidity to fixed acidity. The variable `density_alcohol_ratio` captures the interaction between density and alcohol content. Finally, `quality_score` combines expert rating (`stars`) and marketing appeal (`labelappeal`) to reflect an overall perceived quality of the wine.

In addition, the `stars` variable was transformed into categorical buckets (low, medium, high) to capture potential non-linear effects. These buckets were then converted into dummy variables for use in regression models.

Finally, the cleaned training and evaluation datasets were stored as `wine_train_prep` and `wine_eval_prep` so they can be used directly in the modeling section.

```{r}
#| label: data-preparation
#| echo: true
#| eval: true

# Remove INDEX variable and clean column names
wine_train_prep <- wine_train |> 
  clean_names() |> 
  dplyr::select(-any_of("index"))

wine_eval_prep <- wine_eval |> 
  clean_names() |> 
  dplyr::select(-any_of("index"))

# Identify numeric predictors
numeric_vars <- wine_train_prep |>
  dplyr::select(where(is.numeric)) |>
  dplyr::select(-target) |>
  names()

# 1. Create missingness flags (MUST be done before we overwrite the NAs)
wine_train_prep <- wine_train_prep |>
  mutate(across(
    all_of(numeric_vars),
    ~ ifelse(is.na(.), 1, 0),
    .names = "{.col}_missing"
  ))

wine_eval_prep <- wine_eval_prep |>
  mutate(across(
    all_of(numeric_vars),
    ~ ifelse(is.na(.), 1, 0),
    .names = "{.col}_missing"
  ))

# 2. MEDIAN IMPUTATION (Moved up!)
# We impute the base variables first, so that when we engineer new features
# in the next step, there are no NAs to break the math.
for (v in numeric_vars) {
  med_val <- median(wine_train_prep[[v]], na.rm = TRUE)
  
  wine_train_prep[[v]][is.na(wine_train_prep[[v]])] <- med_val
  wine_eval_prep[[v]][is.na(wine_eval_prep[[v]])] <- med_val
}

# 3. Calculate the minimums from the TRAINING data (now that data is imputed)
min_alc <- min(wine_train_prep$alcohol, na.rm = TRUE)
min_sugar <- min(wine_train_prep$residual_sugar, na.rm = TRUE)
min_fsd <- min(wine_train_prep$free_sulfur_dioxide, na.rm = TRUE)
min_tsd <- min(wine_train_prep$total_sulfur_dioxide, na.rm = TRUE)

# 4. Transformations and feature engineering
wine_train_prep <- wine_train_prep |>
  mutate(
    # Shift values by subtracting the minimum before applying log1p
    log_alcohol = log1p(alcohol - min_alc),
    log_residual_sugar = log1p(residual_sugar - min_sugar),
    log_free_sulfur_dioxide = log1p(free_sulfur_dioxide - min_fsd),
    log_total_sulfur_dioxide = log1p(total_sulfur_dioxide - min_tsd),
    
    sulfur_ratio = total_sulfur_dioxide / ifelse(free_sulfur_dioxide + 1 == 0, 0.001, free_sulfur_dioxide + 1),
    acidity_ratio = volatile_acidity / ifelse(fixed_acidity + 1 == 0, 0.001, fixed_acidity + 1),
    density_alcohol_ratio = density / ifelse(alcohol + 1 == 0, 0.001, alcohol + 1),
    quality_score = stars * label_appeal,
    
    stars_bucket = case_when(
      stars <= 2 ~ "low",
      stars == 3 ~ "medium",
      stars >= 4 ~ "high",
      TRUE ~ "unknown"
    )
  )

wine_eval_prep <- wine_eval_prep |>
  mutate(
    # Apply the exact same training minimums to the evaluation set
    log_alcohol = log1p(alcohol - min_alc),
    log_residual_sugar = log1p(residual_sugar - min_sugar),
    log_free_sulfur_dioxide = log1p(free_sulfur_dioxide - min_fsd),
    log_total_sulfur_dioxide = log1p(total_sulfur_dioxide - min_tsd),
    
    sulfur_ratio = total_sulfur_dioxide / (free_sulfur_dioxide + 1),
    acidity_ratio = volatile_acidity / (fixed_acidity + 1),
    density_alcohol_ratio = density / (alcohol + 1),
    quality_score = stars * label_appeal,
    
    stars_bucket = case_when(
      stars <= 2 ~ "low",
      stars == 3 ~ "medium",
      stars >= 4 ~ "high",
      TRUE ~ "unknown"
    )
  )

# 5. Convert categorical bucket to dummy variables
wine_train_prep <- fastDummies::dummy_cols(
  wine_train_prep,
  select_columns = "stars_bucket",
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

wine_eval_prep <- fastDummies::dummy_cols(
  wine_eval_prep,
  select_columns = "stars_bucket",
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

# 6. Check for remaining missing values in training data
wine_train_prep |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") |>
  filter(Missing > 0)

# 7. Check for remaining missing values in evaluation data
wine_eval_prep |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") |>
  filter(Missing > 0)
```


# Assumptions for Count Regression Model 

Count regression models rest on a set of assumptions that must be evaluated before selecting a final modeling approach. In this section, we assess the key assumptions relevant to Poisson regression, Negative Binomial regression, and their zero-inflated variants using the prepared training data.

## Independence of Observations

Count regression models assume that the outcome for each observation is independent of all others. In the wine dataset, each row represents a distinct wine product, and there is no explicit grouping structure (e.g., repeated measures on the same wine or hierarchical nesting by producer) that would induce correlation between observations. We therefore treat the independence assumption as satisfied.

## Distribution of the Target Variable

The target variable `TARGET` represents the number of wine cases purchased — a non-negative integer count. For Poisson regression, the response is assumed to follow a Poisson distribution, which requires the conditional mean to equal the conditional variance (equidispersion). For Negative Binomial regression, the variance is allowed to exceed the mean (overdispersion), making it more flexible for real-world count data.

```{r}
#| label: tbl-dispersion
#| tbl-cap: "Mean, variance, and dispersion ratio of the TARGET variable"
#| echo: true

dispersion_check <- wine_train_prep |>
  summarise(
    Mean    = mean(target, na.rm = TRUE),
    Variance = var(target, na.rm = TRUE),
    Dispersion_Ratio = var(target, na.rm = TRUE) / mean(target, na.rm = TRUE)
  )

knitr::kable(dispersion_check, digits = 3)
```

A dispersion ratio substantially greater than 1 indicates overdispersion, which would violate the Poisson equidispersion assumption and favor a Negative Binomial specification instead. The dispersion ratio of 1.225 indicates mild overdispersion in the TARGET variable — the variance (3.711) exceeds the mean (3.029), which violates the Poisson assumption of equidispersion. While the  is not extreme, it is sufficient to suggest that a Negative Binomial model may provide a better fit than a standard Poisson model, as it  accounts for this excess variation. We will evaluate both specifications in the modeling section and select the one that minimizes model deviance and information criteria.

## Excess Zeros

Zero-inflated models (ZIP or ZINB) are appropriate when there is a higher frequency of zero counts than would be expected under a Poisson or Negative Binomial distribution alone. As noted in the Data Exploration section, `TARGET` exhibits a high concentration of zero values. We formally examine this by comparing the observed proportion of zeros against the theoretical proportion expected under a fitted Poisson model.


```{r}
#| label: tbl-zero-inflation
#| tbl-cap: "Observed vs. Poisson-expected proportion of zero counts in TARGET"
#| echo: true

lambda_hat <- mean(wine_train_prep$target, na.rm = TRUE)

zero_check <- tibble(
  `Observed % Zeros`  = round(mean(wine_train_prep$target == 0, na.rm = TRUE) * 100, 2),
  `Expected % Zeros (Poisson)` = round(exp(-lambda_hat) * 100, 2)
)

knitr::kable(zero_check)
```

If the observed proportion of zeros substantially exceeds the Poisson expectation, a zero-inflated model will be required to separately model the excess zero-generating process. The observed zero proportion (21.37%) far exceeds what a Poisson model would predict (4.84%), confirming significant excess zeros in `TARGET`. This suggests that a zero-inflated model such as ZIP or ZINB is likely needed to separately account for the portion of wines that generate no sales regardless of their chemical properties.



## Log-Linearity

Count regression models with a log link assume a log-linear relationship between the predictors and the expected count. That is, the log of the expected `TARGET` should be approximately linear in each continuous predictor. We check this visually for the predictors most correlated with `TARGET` — `STARS`, `LabelAppeal`, `AcidIndex`, and `Alcohol` — by plotting the mean log(TARGET + 1) across binned predictor values.

```{r}
#| label: fig-log-linearity
#| fig-cap: "Mean log(TARGET + 1) across binned values of key predictors, used to assess the log-linearity assumption."
#| fig-width: 10
#| fig-height: 6
#| warning: false
#| echo: true

key_vars <- c("stars", "label_appeal", "acid_index", "alcohol")

wine_train_prep |>
  mutate(log_target = log1p(target)) |>
  dplyr::select(log_target, all_of(key_vars)) |>
  pivot_longer(cols = all_of(key_vars), names_to = "variable", values_to = "value") |>
  mutate(bin = cut(value, breaks = 10)) |>
  group_by(variable, bin) |>
  summarise(mean_log_target = mean(log_target, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = bin, y = mean_log_target, group = variable)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(x = "Binned Predictor Value", y = "Mean log(TARGET + 1)")
```

Approximate linearity in each panel supports the log-link assumption. Substantial curvature in any panel would suggest that a transformation or non-linear term for that predictor may be warranted. The log-linearity plots show mixed results across the four key predictors. `label_appeal` displays the clearest log-linear relationship, with mean log(TARGET + 1) rising steadily as label appeal increases, strongly supporting the log-link assumption for this variable. `alcohol` also shows an approximately linear trend with a gradual upward slope, though with minor fluctuations. `stars` shows a positive relationship but with only two visible data points, suggesting limited variation in this discrete variable. `acid_index`, however, shows a non-linear pattern — mean log(TARGET + 1) rises initially before dropping sharply at higher acidity levels — indicating that the log-linearity assumption may not hold for this variable and that a transformation or non-linear term for `acid_index` may improve model fit.


## Multicollinearity

Count regression models, like all generalized linear models, are sensitive to high multicollinearity among predictors. As noted in the correlation matrix (@fig-correlation-matrix), `STARS` and `LabelAppeal` are both positively correlated with `TARGET`, and there is moderate correlation between the acidity variables. We compute Variance Inflation Factors (VIF) on a baseline Poisson model to formally assess multicollinearity.

```{r}
#| label: tbl-vif
#| tbl-cap: "Variance Inflation Factors (VIF) for key predictors in a baseline Poisson model"
#| warning: false
#| echo: true
baseline_formula <- target ~ stars + label_appeal + acid_index + alcohol +
                             volatile_acidity + total_sulfur_dioxide +
                             residual_sugar + chlorides + density + p_h

baseline_poisson <- glm(baseline_formula, data = wine_train_prep, family = poisson(link = "log"))

vif_vals <- vif(baseline_poisson) |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  rename(VIF = 2) |>
  arrange(desc(VIF))

knitr::kable(vif_vals, digits = 3)
```

VIF values exceeding 5 (or conservatively 10) indicate problematic multicollinearity. Variables with high VIF may need to be excluded or combined in the final model specification. All VIF values are extremely close to 1, indicating virtually no multicollinearity among the predictors in the baseline Poisson model. No variable exceeds the conservative threshold of 5, confirming that each predictor contributes independent information to the model. This means multicollinearity will not be a concern during model building, and all variables can be retained without risk of inflated standard errors or unstable coefficient estimates.

## Goodness of Fit

Finally, we assess the overall fit of the baseline Poisson model using the residual deviance relative to its degrees of freedom. A residual deviance substantially larger than the degrees of freedom is a further indicator of overdispersion or model misfit, and may motivate the use of a Negative Binomial or zero-inflated specification.

```{r}
#| label: tbl-deviance
#| tbl-cap: "Residual deviance and degrees of freedom for the baseline Poisson model"
#| echo: true

deviance_check <- tibble(
  `Residual Deviance` = round(deviance(baseline_poisson), 2),
  `Degrees of Freedom` = df.residual(baseline_poisson),
  `Deviance / df`     = round(deviance(baseline_poisson) / df.residual(baseline_poisson), 3)
)

knitr::kable(deviance_check)
```

A deviance-to-df ratio near 1.0 is consistent with a well-fitting Poisson model. Ratios substantially above 1.0 confirm overdispersion and support transitioning to a Negative Binomial or zero-inflated model in the modeling section. The deviance-to-df ratio of **1.443** confirms overdispersion in the baseline Poisson model — the residual deviance (18,452.11) substantially exceeds the degrees of freedom (12,784). This is consistent with our earlier finding that the variance exceeds the mean, and reinforces that a standard Poisson model is not the most appropriate fit for this data. A Negative Binomial or zero-inflated specification will be better suited to account for this excess variation in the modeling section.


# Build models

## Modeling Plan

We follow a workflow centered on building valid models:

1. Partitioning: We split data into a training set for model development and a test set for assessing predictive ability.

2. Poisson GLM: Because our target variable is a count of cases purchased, we utilize Poisson regression. We will use the Pearson dispersion check to verify the model's validity and check for overdispersion.

## Data Splits

```{r}
#| label: split-data
#| eval: true

# Set a seed for reproducibility so you get the same split every time
set.seed(19940211)

# Calculate how many rows make up 80% of the dataset
train_size <- floor(0.80 * nrow(wine_train_prep))

# Randomly select row numbers for the training set
train_indices <- sample(seq_len(nrow(wine_train_prep)), size = train_size)

# Subset the dataset based on those row numbers
wine_train_split <- wine_train_prep[train_indices, ]
wine_test_split <- wine_train_prep[-train_indices, ]
```


## Model 1. Full Model (All untransformed variables)

For our baseline model, we include all available predictor variables in their original, untransformed state. Because our `target` variable represents a count (number of cases purchased), we use a Poisson Generalized Linear Model (`family = "poisson"`).

This full model gives us a starting point to assess the baseline residual deviance, identify which predictors show initial statistical significance, and check for potential overdispersion before we attempt dimensionality reduction.

```{r}
#| label: build-model1
#| eval: true

model1 <- glm(
    formula = target ~ .,
    family = "poisson",
    data = wine_train_split
)

summary(model1)
```

## Model 2. Reduce multicollinearity

Our dataset contains several engineered variables that are direct mathematical combinations of base variables (e.g., `quality_score` is the product of `stars` and `label_appeal`; `sulfur_ratio` is derived from total and free sulfur dioxide). Including both the original base variables and their direct mathematical combinations can cause severe multicollinearity, which artificially inflates the standard errors of our coefficients.

In Model 2, we remove these highly redundant engineered interactions to stabilize our parameter estimates while keeping the primary effects intact.

```{r}
#| label: build-model2
#| eval: true

# Fit Poisson model excluding highly collinear engineered features
model2 <- glm(
    formula = target ~ . - quality_score - sulfur_ratio - acidity_ratio - density_alcohol_ratio,
    family = "poisson",
    data = wine_train_split
)

summary(model2)
```

## Model 3. Negative Binomial Model

In our Assumptions section, we calculated a dispersion ratio of 1.225, indicating that the variance of target slightly exceeds its mean. While a standard Poisson model assumes the variance equals the mean, a Negative Binomial model includes an extra parameter to explicitly handle this overdispersion.

For Model 3, we use the glm.nb() function from the MASS package. We will apply the same reduced feature set used in Model 2 to ensure the model remains stable and free from severe multicollinearity while we test this new distribution.

```{r}
#| label: build-model3
#| eval: true

# Fit a Negative Binomial model to account for mild overdispersion
model3 <- MASS::glm.nb(
    formula = target ~ . - quality_score - sulfur_ratio - acidity_ratio - density_alcohol_ratio,
    data = wine_train_split
)

summary(model3)
```

## Model 4. Backward Stepwise Selection

To balance model complexity and goodness-of-fit, we perform backward stepwise selection starting from our Negative Binomial model (`model3`). This algorithm evaluates the model by iteratively removing the least significant predictors based on the Akaike Information Criterion (AIC). It continues this process until it finds the optimal sub-model where removing any more variables would hurt the model's predictive power more than it would help its simplicity.

```{r}
#| label: build-model4

# Perform backward stepwise selection 
# (trace = 0 hides the massive console output during the iterative process)
model4 <- step(model3, direction = "backward", trace = 0)

summary(model4)
```

# Select Models

```{r}
#| label: model-comparison-table
#| echo: true
#| results: asis
#| eval: false

# Creating a custom helper function to extract all the required metrics
# explicitly evaluating on the provided training dataset.
get_logistic_metrics <- function(model, model_name, data, target_var = "target") {
    # 1. Get predicted probabilities and classes
    probs <- predict(model, newdata = data, type = "response")
    preds <- factor(ifelse(probs > 0.5, 1, 0), levels = c(0, 1))
    actuals <- factor(data[[target_var]], levels = c(0, 1))

    # 2. Generate Confusion Matrix and derived metrics
    cm <- confusionMatrix(preds, actuals, positive = "1")

    # 3. Calculate AUC
    roc_obj <- roc(data[[target_var]], probs, quiet = TRUE)
    auc_val <- as.numeric(pROC::auc(roc_obj))

    # 4. Return results as a tibble
    tibble(
        Model = model_name,
        Deviance = model$deviance,
        Accuracy = cm$overall["Accuracy"],
        Error_Rate = 1 - cm$overall["Accuracy"],
        Precision = cm$byClass["Precision"],
        Sensitivity = cm$byClass["Sensitivity"], # Also Recall
        Specificity = cm$byClass["Specificity"],
        F1 = cm$byClass["F1"],
        AUC = auc_val,
        AIC = AIC(model),
        BIC = BIC(model)
    )
}

models_list <- list(
    "All Untransformed Variables" = model1,
    "Reducing Multicollinearity" = model2,
    "Full Model with Backward Selection" = model_full
)

# Apply the function to all three models, explicitly passing the training split
model_comparison <- imap_dfr(models_list, ~ {
    get_logistic_metrics(.x, .y, crime_test_split)
}) |>
    arrange(`Deviance`)

# Transpose the data and convert to a data frame
model_t <- as.data.frame(t(model_comparison))

# 3. Display the table
knitr::kable(
    model_t,
    digits = 4,
    caption = "Evaluation of Candidate Models on Training Dataset"
)

best_model_info <- model_comparison |> slice(1)

best_model_obj <- models_list[[best_model_info$Model]]
```

For the logistic models, we decided to use the deviance as our primary criterion on the training dataset, which was attained by the `best_model_info$Model`. We've chosen deviance as our primary metric because it directly measures the goodness of fit for logistic regression models, with lower deviance indicating a better fit to the training data. Given our model results, each model had a fairly similar accuracy with our `best_model_info$Model` having the highest accuracy of `round(best_model_info$Accuracy, 3)`. It also had the lowest error rate of `round(best_model_info$Error_Rate, 3)` and was comparable with the other top models in sensitivity and specificity. Overall, this model had the best balance of fit and predictive performance on the training data, which is why we selected it for further evaluation on the test set and for generating predictions on the evaluation set.

Given that our target variable was fairly balanced (49% of all observations had a target value of 1), and our accuracy, Precision, and Sensitivity show that we greatly outperform a random classifier, we are confident that our model is performing well.

```{r}
#| label: stepwise-model-summary
#| echo: true
#| eval: false
# Display the summary of the selected Stepwise model to interpret coefficients and significance
summary(best_model_obj)
```

To make inferences from our top performing model, we can look at the significance of the predictors. The logistic regression model identifies several key environmental and structural factors that significantly predict whether a suburb's crime rate will exceed the median. The most dominant predictor is the concentration of nitrogen oxides (nox), where higher levels are strongly associated with a higher probability of the area being high-crime (z=4.670,p\<0.001). Additionally, the pupil-teacher ratio (ptratio) and the proportion of older homes (age) built before 1940 are both significant positive predictors, suggesting that areas with crowded schools and aging infrastructure are more likely to see elevated crime rates. The model also highlights that as the distance to employment centers (log_dis) increases, so does the likelihood of high crime, along with a significant effect from the ratio of rooms to lower-status population (rm_lstat_ratio). Interestingly, many traditional economic indicators, such as median home values (medv), property tax rates (log_tax), and zoning for large lots (log_zn), did not reach statistical significance, indicating that in this specific dataset, industrial and demographic markers are more reliable indicators of crime than general wealth or proximity to the Charles River.

# Results

```{r}
#| label: predictions-on-eval-set
#| echo: true
#| eval: false
# Generate predictions on the evaluation dataset using the selected model
eval_predictions <- predict(best_model_obj, newdata = crime_eval_clean, type = "response")

crime_eval_results <- crime_eval_clean |>
    mutate(
        # If probability > 0.5, it's a 1 (high crime), otherwise 0
        PREDICTED_TARGET = ifelse(eval_predictions > 0.5, 1, 0)
    )

# Write predictions to a csv
write_csv(crime_eval_results, "crime_evaluation_predictions.csv")

# Display the first 10 predictions in the rendered PDF
knitr::kable(
    head(crime_eval_results |> select(PREDICTED_TARGET), 10),
    caption = "Excerpts of Stepwise Model Win Predictions for Evaluation Set"
)
```

Given our evaluation set and our stepwise model, we are able to create predictions on the evaluation set on which neighborhoods are classified as high-crime or low-crime. The first 10 predictions are shown in the table above. We write these predictions to a CSV file for submission.

# Conclusion

In conclusion, our `best_model_info$Model` model, which was selected based on deviance and other performance metrics, demonstrates strong predictive capabilities for classifying neighborhoods as high-crime or low-crime based on a variety of environmental and structural factors. The significant predictors identified in the model provide valuable insights into the underlying drivers of crime in urban areas, highlighting the importance of air pollution, school crowding, aging infrastructure, and proximity to employment centers. These findings can inform policymakers and community leaders in their efforts to target interventions and allocate resources effectively to reduce crime rates. Future work could explore additional modeling techniques, such as regularization or ensemble methods, to further enhance predictive performance and interpretability.
