---
title: "Insurance"
subtitle: "Homework #4"
format: pdf
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong
  - name: Said Naqwe
date: April 26, 2026
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Nana Frimpong, Said Naqwe}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: load-packages
#| include: false

# Load necessary libraries
library(tidyverse)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
library(caret)
library(pROC)
library(fastDummies)
library(tidytable)
library(car)
```

# Introduction

We model two outcomes in a auto insurance outcome variables: (1) whether the person was in a car crash (binary) and (2) the amount of money it will cost if the person does crash their car (continuous). Our objective is to build multiple linear regression and binary logistic regression models on the training data to predict the probability that a person will crash their car and also predict the cost of the crash if it occurs.

# Data Exploration and Cleaning

```{r}
#| label: data-exploration
#| warning: false

# Load the training and evaluation datasets
insurance_train <- read_csv("./data/insurance_training_data.csv")
insurance_eval <- read_csv("./data/insurance-evaluation-data.csv")

# Count rows and columns
nrows <- insurance_train |> nrow()
ncols <- insurance_train |> length()

# Get missingness
perc_complete <- round(insurance_train |> n_complete() / nrows / ncols * 100, 2)
```

The insurance training data set contains `r nrows` rows and `r ncols` features, and is `r perc_complete` percent complete. We will need to address the missingness in the data preparation section.

```{r}
#| label: preview-categorical-variables
#| include: false

# Get list of categorical variables
categorical_vars <- insurance_train |> select(where(is.character)) |> names()

# Preview the categorical variables
insurance_train |> select(where(is.character)) |> head()
```

There are several variables in the dataset coded as characters. There are several currency variables (`INCOME`, `HOME_VAL`, `BLUEBOOK`, `OLDCLAIM`) that should be changed to numeric variables. The rest of the character variables should be encoded as dummy variables for modeling in the following chunk. We also rename some variables for clarity. We also remove the `index` variable as it is not a predictor and will not be used in modeling.

```{r}
#| label: fix-categorical-variables
#| warning: false
# Convert currency variables to numeric by removing "$" and converting to numeric
currency_vars <- c("INCOME", "HOME_VAL", "BLUEBOOK", "OLDCLAIM")

# Character columns to one-hot encode (exclude currency columns)
char_vars <- insurance_train |>
    select(where(is.character)) |>
    names() |>
    setdiff(currency_vars)

insurance_train_dummies <- insurance_train |>
    mutate(
        # Convert currency variables to numeric
        across(all_of(currency_vars), parse_number)
    ) |>
    fastDummies::dummy_cols(
        select_columns = char_vars,
        remove_selected_columns = TRUE,
        remove_first_dummy = TRUE
    ) |> 
    clean_names() |> 
    rename(
        female = sex_z_f,
        education_phd = education_ph_d,
        education_high_school = education_z_high_school,
        car_type_suv = car_type_z_suv,
        job_blue_collar = job_z_blue_collar,
        urbanicity_highly_rural = urbanicity_z_highly_rural_rural
    ) |> 
    select(-index)

insurance_eval_dummies <- insurance_eval |>
    mutate(
        across(all_of(currency_vars), parse_number)
    ) |>
    fastDummies::dummy_cols(
        select_columns = char_vars,
        remove_selected_columns = TRUE,
        remove_first_dummy = TRUE
    ) |> 
    clean_names()|> 
    rename(
        female = sex_z_f,
        education_phd = education_ph_d,
        education_high_school = education_z_high_school,
        car_type_suv = car_type_z_suv,
        job_blue_collar = job_z_blue_collar,
        urbanicity_highly_rural = urbanicity_z_highly_rural_rural
    ) |> 
    select(-index)

```

In previewing summary statistics for all variables, some of the minimum values appear to be in error. For example, `car_age` has a negative value, and there are a high number of 0 values in `car_age`, `home_val`, which may be data entry errors--we will treat them as missing and replace them with imputed values.

```{r}
#| label: preview-summ-stats
#| output: false

insurance_train_dummies |> summary()

insurance_train_dummies <- insurance_train_dummies |>
    mutate(
        car_age = ifelse(car_age <= 0, NA, car_age),
        home_val = ifelse(home_val == 0, NA, home_val)
    )
```

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for all variables in the training dataset"
#| echo: false

tbl_summ_stats <- insurance_train_dummies |>
    skim_without_charts() |>
    janitor::clean_names() |>
    as_tibble() |>
    rename(Variable = skim_variable, "Completeness rate" = complete_rate, "Mean" = numeric_mean, "SD" = numeric_sd, "P50" = numeric_p50, "Min" = numeric_p0, "P25" = numeric_p25, "P75" = numeric_p75, "Max" = numeric_p100) |>
    select(Variable, "Completeness rate", Mean, Min, P25, P50, P75, Max) |>
    mutate(
        across(where(is.numeric), ~ signif(., 2)),
        Mean = format(Mean, scientific = FALSE, trim = TRUE)
    ) |>
    knitr::kable()

tbl_summ_stats
```

@tbl-summary-statistics presents descriptive characteristics of the training data. The target variables are `target_flag`, which represents whether the person was in a car crash (1) or not (0), and `target_amt`, which represents the cost of the crash if it occurs. The other variables represent various information on auto insurance for different individuals. 

```{r}
#| label: fig-correlation-matrix
#| fig-cap: "Correlation matrix of all variables in the training dataset. Note: Correlations with absolute value less than 0.2 are suppressed."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Get correlation matrix
insurance_train_cor <- insurance_train_dummies |>
    select(where(is.numeric)) |> 
    cor(use = "pairwise.complete.obs")

insurance_train_cor_plot <- insurance_train_cor
insurance_train_cor_plot[abs(insurance_train_cor_plot) < 0.2] <- NA

corrplot(insurance_train_cor_plot, method = "circle", type = "upper", sig.level = 0.05, insig = "blank", addCoef.col = "black", number.cex = 0.7, tl.cex = 0.7, tl.srt = 45, na.label = " ")

most_corr_w_target_flag <- insurance_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "target_flag", Variable != "target_amt") |>
    transmute(
        Variable,
        Correlation = target_flag,
        Abs_Correlation = abs(target_flag)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.2) |> pull(Variable)

most_corr_w_target_amt <- insurance_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "target_flag", Variable != "target_amt") |>
    transmute(
        Variable,
        Correlation = target_amt,
        Abs_Correlation = abs(target_amt)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.2) |> 
    pull(Variable)
```

The correlation between all variables are shown @fig-correlation-matrix. The variables most correlated with `target_flag`, whether there was a car crash, (besides the amount the crash costed) are `r paste(most_corr_w_target_flag, collapse = ", ")`. This means that people who work or live in a more rural area are less likely to be in a crash, and people who  have more motor vehicle record points or have more claims in the past 5 years are more likely to be in a crash. There were no predictor variables strongly correlated with `target_amt`, the cost if the crash occurs. 

Lastly, for continuous variables, we also looked at the distribution of each variable to check for skewness and outliers. 

```{r}
#| label: fig-distributions
#| fig-cap: "Distribution of continuous variables in the training dataset."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Keep only continuous
continuous_vars <- insurance_train_dummies |>
    select(where(is.numeric)) |>
    select(where(~ dplyr::n_distinct(stats::na.omit(.x)) > 10)) |>
    names()

insurance_train_dummies |>
    select(all_of(continuous_vars)) |>
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    facet_wrap(~ variable, scales = "free", ncol = 4) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    labs(x = NULL, y = "Count")

```

There are many 1-year old cars in the dataset, showing a high spike at 1 in the `car_age` variable. Dollar amounts such as `bluebook`, `home_val`, `income`, and `target_amt` are right-skewed, which is expected for financial variables. The same goes for travel time (`travtime`). Most people have no motor vehicle record points (`mvr_pts`) or claims in the past 5 years (`oldclaim`), which is expected for a general population. Years on the job (`yoj`) is almost normally distributed, but has a spike at 0, which may represent people who are unemployed or have been on the job for less than a year.

# Data Preparation

The goal of this section is to prepare the insurance training and evaluation data for modeling. The assignment requires transformations of the original variables or the creation of new variables when useful. In this dataset, preparation included missing value treatment, missingness flags, mathematical transformations, and derived variables.

```{r}
#| label: data-preparation
#| warning: false

# Identify numeric predictor variables
numeric_vars <- insurance_train_dummies |>
  select(where(is.numeric)) |>
  select(-target_flag, -target_amt) |>
  names()

# Create missingness flags before imputation
insurance_train_prepared <- insurance_train_dummies |>
  mutate(
    age_missing = ifelse(is.na(age), 1, 0),
    car_age_missing = ifelse(is.na(car_age), 1, 0),
    home_val_missing = ifelse(is.na(home_val), 1, 0),
    income_missing = ifelse(is.na(income), 1, 0),
    yoj_missing = ifelse(is.na(yoj), 1, 0)
  )

insurance_eval_prepared <- insurance_eval_dummies |>
  mutate(
    age_missing = ifelse(is.na(age), 1, 0),
    car_age_missing = ifelse(is.na(car_age), 1, 0),
    home_val_missing = ifelse(is.na(home_val), 1, 0),
    income_missing = ifelse(is.na(income), 1, 0),
    yoj_missing = ifelse(is.na(yoj), 1, 0)
  )

# Median imputation using training medians
for (v in numeric_vars) {
  med_val <- median(insurance_train_prepared[[v]], na.rm = TRUE)

  insurance_train_prepared[[v]][is.na(insurance_train_prepared[[v]])] <- med_val
  insurance_eval_prepared[[v]][is.na(insurance_eval_prepared[[v]])] <- med_val
}

# Create transformed and engineered variables
insurance_train_prepared <- insurance_train_prepared |>
  mutate(
    log_income = log1p(income),
    log_home_val = log1p(home_val),
    log_bluebook = log1p(bluebook),
    log_oldclaim = log1p(oldclaim),
    log_travtime = log1p(travtime),

    claim_rate = oldclaim / (clm_freq + 1),
    car_value_age_ratio = bluebook / (car_age + 1),
    risk_points_claims = mvr_pts + clm_freq,

    age_bucket = case_when(
      age < 25 ~ "young_driver",
      age >= 25 & age < 60 ~ "adult_driver",
      age >= 60 ~ "older_driver",
      TRUE ~ "unknown"
    )
  )

insurance_eval_prepared <- insurance_eval_prepared |>
  mutate(
    log_income = log1p(income),
    log_home_val = log1p(home_val),
    log_bluebook = log1p(bluebook),
    log_oldclaim = log1p(oldclaim),
    log_travtime = log1p(travtime),

    claim_rate = oldclaim / (clm_freq + 1),
    car_value_age_ratio = bluebook / (car_age + 1),
    risk_points_claims = mvr_pts + clm_freq,

    age_bucket = case_when(
      age < 25 ~ "young_driver",
      age >= 25 & age < 60 ~ "adult_driver",
      age >= 60 ~ "older_driver",
      TRUE ~ "unknown"
    )
  )

# Convert age bucket into dummy variables
insurance_train_prepared <- fastDummies::dummy_cols(
  insurance_train_prepared,
  select_columns = "age_bucket",
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

insurance_eval_prepared <- fastDummies::dummy_cols(
  insurance_eval_prepared,
  select_columns = "age_bucket",
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

# Confirm no missing values remain
sum(is.na(insurance_train_prepared))
sum(is.na(insurance_eval_prepared))
```

## Data Splits

First, we will split our dataset into a training and testing set (80:20 ratio). We will use the training set to build our models and the testing set to evaluate their performance. The training set will be used to fit the models, while the testing set will be used to assess how well the models generalize to new data.

```{r}
#| label: split-data

set.seed(19940211)
train_indices <- sample(nrow(insurance_train_prepared), size = 0.8 * nrow(insurance_train_prepared))
insurance_train_split <- insurance_train_prepared[train_indices, ]
insurance_test_split <- insurance_train_prepared[-train_indices, ]
```

# Check assumptions for Binary logistic regression
1. Binary Outcome
```{r}
# Check assumptions for binary logistic regression

#| label: assumption-binary-outcome
#| echo: true

# Confirming that the target is binary
insurance_train_split |>
  count(target_flag) |>
  mutate(prop = n / sum(n))
```

The response variable target_flag is confirmed to be binary, taking two values (0: no car crash) and (1: car crash occurred). In the training split, approximately 26% of policyholders were involved in a crash and 74% were not. Unlike the crime dataset in Homework 3, there is a moderate class imbalance here, which means accuracy alone may be a misleading performance metric. We will rely on sensitivity, specificity, and AUC-ROC when evaluating our models.

2. Independence of Observations

\vspace{0.5em}

Each record in the dataset represents a distinct policyholder, and there is no repeated-measures or clustered structure in the data. Therefore, the independence of observations assumption is satisfied by design and requires no further adjustment.

3. No Multicollinearity (VIF Check)

```{r}
#| label: assumption-multicollinearity
#| echo: true
#| warning: false

library(car)

# Fit a reference logistic model with core untransformed predictors
model_ref <- glm(
  target_flag ~ age + homekids + yoj + income + home_val + travtime +
                bluebook + tif + oldclaim + clm_freq + mvr_pts + car_age,
  data = insurance_train_split,
  family = binomial
)

# Compute VIF
vif_vals <- vif(model_ref) |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  rename(VIF = 2) |>
  arrange(desc(VIF))

vif_vals |> knitr::kable(caption = "Variance Inflation Factors for Reference Model")
```
 

VIF values were calculated for all predictors to check for multicollinearity. 
A VIF above 10 is considered to be severe, while between 5 and 10 should warrant 
caution. None of the variables exceed the threshold of 10, indicating multicollinearity 
is not present. The highest VIF belongs to income (VIF = 2.45), followed by home_val 
(VIF = 1.99). All remaining predictors fall well below 5 and are at acceptable levels 
of multicollinearity. Therefore, no variables need to be removed on the basis of 
multicollinearity, and all predictors may be considered for model building.

4. Linearity of Log-Odds (Box-Tidwell Test)

```{r}
#| label: assumption-linearity-log-odds
#| echo: true
#| warning: false

# Box-Tidwell test: add interaction of each continuous predictor with its log
# to test whether the log-odds relationship is linear
continuous_vars <- c("age", "homekids", "yoj", "travtime", "bluebook",
                     "tif", "oldclaim", "clm_freq", "mvr_pts", "car_age")

# Create interaction terms x * log(x) for each continuous predictor
# (use log1p to handle zeros)
insurance_bt <- insurance_train_split |>
  mutate(across(all_of(continuous_vars), ~ . * log1p(.), .names = "{.col}_logx"))

bt_formula <- as.formula(paste(
  "target_flag ~",
  paste(continuous_vars, collapse = " + "),
  "+",
  paste(paste0(continuous_vars, "_logx"), collapse = " + ")
))

model_bt <- glm(bt_formula, data = insurance_bt, family = binomial)

# Extract p-values for the interaction terms only
bt_results <- summary(model_bt)$coefficients |>
  as.data.frame() |>
  rownames_to_column("Term") |>
  filter(str_detect(Term, "_logx")) |>
  mutate(Variable = str_remove(Term, "_logx"),
         Significant = `Pr(>|z|)` < 0.05) |>
  select(Variable, Estimate, `Pr(>|z|)`, Significant) |>
  mutate(across(where(is.numeric), ~ signif(., 3)))

bt_results |> knitr::kable(caption = "Box-Tidwell Test for Linearity of Log-Odds")
```

The Box-Tidwell test was used to assess whether each continuous predictor has a 
linear relationship with the log-odds of a car crash. A significant p-value (< 0.05) 
for the interaction term indicates a violation of this assumption. Six variables were 
flagged as significant: age (p < 0.001), homekids (p = 0.003), travtime (p = 0.002), 
bluebook (p < 0.001), tif (p = 0.048), and clm_freq (p < 0.001). This indicates that 
these predictors do not have a strictly linear relationship with the log-odds of a car 
crash. Log-transformed versions of travtime and bluebook (log_travtime, log_bluebook) 
were already created in the data preparation step and will be used in model building. 
For age, homekids, tif, and clm_freq, we will consider log1p transformations or 
binning strategies during model building to address the non-linearity. The remaining 
four predictors — yoj, oldclaim, mvr_pts, and car_age — did not show evidence of 
violating the linearity of log-odds assumption and may enter models in their original 
form.

# Check Assumptions for Linear Regression

For the linear regression model, we model `target_amt` only on records where a 
crash occurred (`target_flag == 1`), since a value of zero indicates no crash 
rather than an actual claim cost of zero.

```{r}
#| label: linear-regression-subset

# Subset to only crashed records for linear regression assumption checks
insurance_train_crash <- insurance_train_split |> 
  filter(target_flag == 1)

nrow(insurance_train_crash)
```

1. Linearity

```{r}
#| label: assumption-linearity
#| echo: true
#| warning: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Scatterplots of continuous predictors vs. target_amt"

continuous_vars_lm <- c("age", "yoj", "income", "home_val", "travtime",
                        "bluebook", "tif", "oldclaim", "clm_freq", 
                        "mvr_pts", "car_age")

insurance_train_crash |>
  select(all_of(continuous_vars_lm), target_amt) |>
  pivot_longer(-target_amt, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value, y = target_amt)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(x = NULL, y = "target_amt")
```


The scatterplots show the relationship between each continuous predictor and 
`target_amt`. Most predictors show a weak or near-flat linear relationship with 
the response, which is consistent with the earlier finding that no predictor 
variables were strongly correlated with `target_amt` in the correlation matrix. 
The only exception is `bluebook`, which shows a slight positive trend, suggesting 
that more expensive vehicles are associated with higher crash costs. The heavy 
concentration of points at lower values of `target_amt` across all predictors 
reflects the right-skewed distribution of the response variable noted during data 
exploration. Log transformations of `bluebook`, `income`, `oldclaim`, and 
`travtime` created in the data preparation step will be used during model building 
to address this.

2. Independence of Observations

\vspace{0.5em}

Each record in the dataset represents a distinct policyholder involved in a crash. 
There is no repeated-measures or clustered structure in the data. Therefore, the 
independence of observations assumption is satisfied by design and requires no 
further adjustment.

3. Homoscedasticity (Constant Variance)

```{r}
#| label: assumption-homoscedasticity
#| echo: true
#| warning: false
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Residuals vs. Fitted values plot for homoscedasticity check"

# Fit a reference linear model on crash records only
model_ref_lm <- lm(
  target_amt ~ age + yoj + income + home_val + travtime +
               bluebook + tif + oldclaim + clm_freq + mvr_pts + car_age,
  data = insurance_train_crash
)

# Residuals vs fitted plot
plot(model_ref_lm, which = 1, 
     main = "Residuals vs Fitted",
     pch = 20, col = "steelblue")
```

The residuals vs. fitted plot shows a clear fanning pattern, where the spread 
of residuals increases as fitted values increase. This indicates heteroscedasticity 
— the variance of residuals is not constant across fitted values. Several high-leverage 
outliers (observations 233, 650, and 1372) are visible with residuals exceeding 
$60,000, pulling the fitted line downward at higher values. A log transformation 
of `target_amt` will be considered during model building to stabilize the variance.



4. Normality of Residuals

```{r}
#| label: assumption-normality
#| echo: true
#| warning: false
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Q-Q plot of residuals for normality check"

# Q-Q plot
plot(model_ref_lm, which = 2,
     main = "Normal Q-Q Plot of Residuals",
     pch = 20, col = "steelblue")
```


```{r}
#| label: assumption-normality-shapiro
#| echo: true
#| warning: false

# Shapiro-Wilk test on a sample (max 5000 observations)
set.seed(19940211)
residuals_sample <- residuals(model_ref_lm)

if (length(residuals_sample) > 5000) {
  residuals_sample <- sample(residuals_sample, 5000)
}

shapiro.test(residuals_sample)
```
The Q-Q plot and Shapiro-Wilk test are used to assess whether the residuals are 
normally distributed. Departures from the diagonal line in the Q-Q plot indicate 
non-normality. The Shapiro-Wilk test confirms a violation of this assumption 
(W = 0.479, p < 0.001), indicating that the residuals are significantly non-normal. 
The same three outliers identified in the residuals vs. fitted plot (observations 
233, 650, and 1372) are again visible with standardized residuals exceeding 8. 
A log transformation of `target_amt` will be applied during model building to 
address both the non-normality and heteroscedasticity identified in assumptions 
3 and 4.

# Build models

We will build models for two separate outcomes:

1. **target_flag** (binary): Whether the person was in a car crash (1) or not (0)
2. **target_amt** (continuous): The amount of money the crash cost if it occurs

For each outcome, we compare three models as per the textbook methods: 
- Model 1: Full model with all predictors 
- Model 2: Reduced model (removing highly correlated or insignificant variables) 
- Model 3: Backward stepwise selection model

## Model 1. Full Logistic and Linear Models


```{r}
#| label: build-model1-insurance

# Model 1: Full Logistic Model (target_flag)
# This model uses all available numeric predictors
model1_logistic <- glm(
    formula = target_flag ~ .,
    data = insurance_train_split |> select(-target_amt),
    family = binomial
)

cat("\n=== Full Logistic Model Summary ===\n")
summary(model1_logistic)

# Model 1: Full Linear Model (target_amt)
# This model uses all available numeric predictors
model1_linear <- lm(
    formula = target_amt ~ .,
    data = insurance_train_split |> select(-target_flag)
)

cat("\n=== Full Linear Model Summary ===\n")
summary(model1_linear)
```

As shown above, the most significant predictors for `target_flag` (whether a crash occurred) include variables related to the driver's profile and history. The AIC for this logistic model is `r AIC(model1_logistic)`.

Similarly, for `target_amt` (cost of crash), the full linear model identifies significant predictors. The R-squared for this model is `r summary(model1_linear)$r.squared`.

## Model 2. Reduced Models (Addressing Multicollinearity)

In the Data Exploration section, we identified several variables with high correlations. For the reduced models, we remove highly correlated predictors while retaining theoretically important ones.


```{r}
#| label: build-model2-insurance

# Model 2: Reduced Logistic Model
# Keep only important predictors based on correlation analysis
reduced_vars <- c("kidsdriv", "age", "homekids", "yoj", "income", "home_val",
                  "travtime", "bluebook", "tif", "oldclaim", "clm_freq", "mvr_pts", "car_age")

model2_logistic <- glm(
    formula = as.formula(paste("target_flag ~", paste(reduced_vars, collapse = "+"))),
    data = insurance_train_split,
    family = binomial
)

cat("\n=== Reduced Logistic Model Summary ===\n")
summary(model2_logistic)

# Model 2: Reduced Linear Model
model2_linear <- lm(
    formula = as.formula(paste("target_amt ~", paste(reduced_vars, collapse = "+"))),
    data = insurance_train_split
)

cat("\n=== Reduced Linear Model Summary ===\n")
summary(model2_linear)
```

The AIC for the reduced logistic model is `r AIC(model2_logistic)`, and the R-squared for the reduced linear model is `r summary(model2_linear)$r.squared`.

## Model 3. Backward Stepwise Selection Models

For this model, we performed backward stepwise selection from the full model to identify the most important predictors.

```{r}
#| label: build-model3-insurance

# Model 3: Backward Stepwise Logistic Model
full_logistic <- glm(
    formula = target_flag ~ .,
    data = insurance_train_split |> select(-target_amt),
    family = binomial
)

model3_logistic <- step(full_logistic, direction = "backward", trace = 0)

cat("\n=== Stepwise Logistic Model Summary ===\n")
summary(model3_logistic)

# Model 3: Backward Stepwise Linear Model
full_linear <- lm(
    formula = target_amt ~ .,
    data = insurance_train_split |> select(-target_flag)
)

model3_linear <- step(full_linear, direction = "backward", trace = 0)

cat("\n=== Stepwise Linear Model Summary ===\n")
summary(model3_linear)
```

The AIC for the stepwise logistic model is `r AIC(model3_logistic)`, and the R-squared for the stepwise linear model is `r summary(model3_linear)$r.squared`.

Of these three models, the stepwise models have the lowest AIC for logistic and highest R-squared for linear. In the next section, we will evaluate the performance of these models to determine which ones are best for predicting the insurance outcomes.

# Select Models

We evaluate both logistic and linear regression models using standard metrics from the textbook methods.

## Evaluation Metrics Function

```{r}
#| label: model-evaluation-functions

#' Evaluate Logistic Regression Models
#' @param model A glm logistic model
#' @param model_name Character name of the model
#' @param data Test dataframe
#' @param target_col Name of target column
#' @return Tibble with evaluation metrics
evaluate_logistic_model <- function(model, model_name, data, target_col = "target_flag") {
    probs <- predict(model, newdata = data, type = "response")
    preds <- factor(ifelse(probs > 0.5, 1, 0), levels = c(0, 1))
    actuals <- factor(data[[target_col]], levels = c(0, 1))
    
    cm <- confusionMatrix(preds, actuals, positive = "1")
    roc_obj <- roc(data[[target_col]], probs, quiet = TRUE)
    auc_val <- as.numeric(pROC::auc(roc_obj))
    
    tibble(
        Model = model_name,
        Deviance = model$deviance,
        Accuracy = cm$overall["Accuracy"],
        Error_Rate = 1 - cm$overall["Accuracy"],
        Precision = cm$byClass["Precision"],
        Sensitivity = cm$byClass["Sensitivity"],
        Specificity = cm$byClass["Specificity"],
        F1 = cm$byClass["F1"],
        AUC = auc_val,
        AIC = AIC(model),
        BIC = BIC(model)
    )
}

#' Evaluate Linear Regression Models
#' @param model An lm linear model
#' @param model_name Character name of the model
#' @param data Test dataframe
#' @param target_col Name of target column
#' @return Tibble with evaluation metrics
evaluate_linear_model <- function(model, model_name, data, target_col = "target_amt") {
    preds <- predict(model, newdata = data)
    actuals <- data[[target_col]]
    
    # Calculate metrics
    mse <- mean((actuals - preds)^2)
    rmse <- sqrt(mse)
    mae <- mean(abs(actuals - preds))
    r2 <- summary(model)$r.squared
    adj_r2 <- summary(model)$adj.r.squared
    
    tibble(
        Model = model_name,
        MSE = mse,
        RMSE = rmse,
        MAE = mae,
        R_Squared = r2,
        Adj_R_Squared = adj_r2,
        AIC = AIC(model),
        BIC = BIC(model)
    )
}

# Evaluate all logistic models
cat("=== Evaluating Logistic Regression Models ===\n")
logistic_models <- list(
    "Full Logistic Model" = model1_logistic,
    "Reduced Logistic Model" = model2_logistic,
    "Stepwise Logistic Model" = model3_logistic
)

logistic_comparison <- imap_dfr(logistic_models, ~ {
    evaluate_logistic_model(.x, .y, insurance_test_split)
}) |> arrange(AIC)

# Display logistic model comparison
logistic_comparison |> 
    knitr::kable(digits = 4, caption = "Logistic Model Comparison")

# Select best logistic model
best_logistic <- logistic_comparison |> slice(1)
cat("\nBest Logistic Model:", best_logistic$Model, "\n")

# Evaluate all linear models
cat("\n=== Evaluating Linear Regression Models ===\n")
linear_models <- list(
    "Full Linear Model" = model1_linear,
    "Reduced Linear Model" = model2_linear,
    "Stepwise Linear Model" = model3_linear
)

linear_comparison <- imap_dfr(linear_models, ~ {
    evaluate_linear_model(.x, .y, insurance_test_split)
}) |> arrange(desc(Adj_R_Squared))

# Display linear model comparison
linear_comparison |> 
    knitr::kable(digits = 4, caption = "Linear Model Comparison")

# Select best linear model
best_linear <- linear_comparison |> slice(1)
cat("\nBest Linear Model:", best_linear$Model, "\n")
```

For the logistic models, we use AIC as our primary criterion for model selection, with lower AIC indicating better model fit. The `r best_logistic$Model` achieved the lowest AIC of `r best_logistic$AIC`, with an accuracy of `r round(best_logistic$Accuracy, 3)` and AUC of `r round(best_logistic$AUC, 3)`.

For the linear regression models, we use adjusted R-squared to select the best model, with higher values indicating better model fit. The `r best_linear$Model` achieved the highest adjusted R-squared of `r round(best_linear$Adj_R_Squared, 3)` with RMSE of `r round(best_linear$RMSE, 2)`.

## Selected Model Summaries

```{r}
#| label: selected-model-summaries

# Display summary of best logistic model
cat("=== Best Logistic Model Summary ===\n")
summary(logistic_models[[best_logistic$Model]])

# Display summary of best linear model
cat("\n=== Best Linear Model Summary ===\n")
summary(linear_models[[best_linear$Model]])
```

To make inferences from our top performing models, we can look at the significance of the predictors.

For the logistic regression model (target_flag), the model identifies key factors that significantly predict whether a person will be in a car crash. The most important predictors include variables related to driving history, vehicle characteristics, and demographic factors.

For the linear regression model (target_amt), significant predictors help explain the variation in the cost of claims when crashes occur. The most important predictors include variables related to driving history and vehicle characteristics.

# Results

Generate predictions on the evaluation dataset using the selected models:

```{r}
#| label: predictions-on-eval-set
#| echo: true

# Get the best model names
best_logistic_name <- best_logistic$Model
best_linear_name <- best_linear$Model

# Generate predictions for target_flag (probability of crash)
eval_probs <- predict(
    logistic_models[[best_logistic_name]], 
    newdata = insurance_eval_prepared, 
    type = "response"
)

# Generate predictions for target_amt (cost if crash)
eval_amounts <- predict(
    linear_models[[best_linear_name]], 
    newdata = insurance_eval_prepared
)

# Create results dataframe
insurance_eval_results <- insurance_eval_prepared |>
    mutate(
        PREDICTED_target_flag = ifelse(eval_probs > 0.5, 1, 0),
        PREDICTED_target_amt = ifelse(PREDICTED_target_flag == 1, 
                                       round(eval_amounts, 2), 
                                       0)
    )

# Write predictions to CSV
write_csv(insurance_eval_results, "insurance_evaluation_predictions.csv")

# Display sample predictions
cat("Sample predictions for evaluation data:\n")
insurance_eval_results |>
    select(PREDICTED_target_flag, PREDICTED_target_amt) |>
    head(10) |>
    knitr::kable(caption = "Sample Insurance Predictions")
```

We generate predictions for both outcomes:

1. **target_flag**: Whether the person will be in a car crash (binary: 0 or 1)
2. **target_amt**: The cost if a crash occurs (continuous dollar amount)

The predictions are written to `insurance_evaluation_predictions.csv` for submission.

# Conclusion

In conclusion, our best models for predicting the probability that a person will crash their car and the cost if a crash occurs were the `r best_logistic_name` and `r best_linear_name`, respectively. The best logistic regression model had respectable performance with accuracy of `r round(best_logistic$Accuracy, 3)` and AUC of `r round(best_logistic$AUC, 3)`. However, the adjusted R-squared of the best linear regression model was only `r round(best_linear$Adj_R_Squared, 3)`, meaning that it does not explain the variation in the target variable.

The significant predictors in the logistic regression model suggested that driving history and vehicle characteristics are important determinants of the probability of a crash, which is consistent with our expectations (i.e., if you have a history of reckless driving, you're more likely to crash again). These findings can help insurance companies predict drivers' accident risk, which informs the cost of insurance premiums. Other modeling approaches, such as machine learning, may be needed to develop a better predictive model of the cost of a crash.