---
title: "Insurance"
subtitle: "Homework #4"
format: pdf
author:
  - name: Naomi Buell
  - name: Richie Rivera
  - name: Alexander Simon
  - name: Nana Frimpong
  - name: Said Naqwe
date: April 26, 2026
documentclass: article
classoption: 11pt
geometry:
  - margin=1in
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{setspace}
  - \onehalfspacing
  - \pagestyle{fancy}
  - \lhead{DATA 621}
  - \rhead{Naomi Buell, Richie Rivera, Alexander Simon, Nana Frimpong, Said Naqwe}
  - \setcounter{secnumdepth}{7}
---

```{r}
#| label: load-packages
#| include: false

# Load necessary libraries
library(tidyverse)
library(skimr)
library(corrplot)
library(GGally)
library(janitor)
library(caret)
library(pROC)
library(fastDummies)
library(tidytable)
```

# Introduction

We model two outcomes in a auto insurance outcome variables: (1) whether the person was in a car crash (binary) and (2) the amount of money it will cost if the person does crash their car (continuous). Our objective is to build multiple linear regression and binary logistic regression models on the training data to predict the probability that a person will crash their car and also predict the cost of the crash if it occurs.

# Data Exploration and Cleaning

```{r}
#| label: data-exploration
#| warning: false

# Load the training and evaluation datasets
insurance_train <- read_csv("./data/insurance_training_data.csv")
insurance_eval <- read_csv("./data/insurance-evaluation-data.csv")

# Count rows and columns
nrows <- insurance_train |> nrow()
ncols <- insurance_train |> length()

# Get missingness
perc_complete <- round(insurance_train |> n_complete() / nrows / ncols * 100, 2)
```

The insurance training data set contains `r nrows` rows and `r ncols` features, and is `r perc_complete` percent complete. We will need to address the missingness in the data preparation section.

```{r}
#| label: preview-categorical-variables
#| include: false

# Get list of categorical variables
categorical_vars <- insurance_train |> select(where(is.character)) |> names()

# Preview the categorical variables
insurance_train |> select(where(is.character)) |> head()
```

There are several variables in the dataset coded as characters. There are several currency variables (`INCOME`, `HOME_VAL`, `BLUEBOOK`, `OLDCLAIM`) that should be changed to numeric variables. The rest of the character variables should be encoded as dummy variables for modeling in the following chunk. We also rename some variables for clarity. We also remove the `index` variable as it is not a predictor and will not be used in modeling.

```{r}
#| label: fix-categorical-variables
#| warning: false
# Convert currency variables to numeric by removing "$" and converting to numeric
currency_vars <- c("INCOME", "HOME_VAL", "BLUEBOOK", "OLDCLAIM")

# Character columns to one-hot encode (exclude currency columns)
char_vars <- insurance_train |>
    select(where(is.character)) |>
    names() |>
    setdiff(currency_vars)

insurance_train_dummies <- insurance_train |>
    mutate(
        # Convert currency variables to numeric
        across(all_of(currency_vars), parse_number)
    ) |>
    fastDummies::dummy_cols(
        select_columns = char_vars,
        remove_selected_columns = TRUE,
        remove_first_dummy = TRUE
    ) |> 
    clean_names() |> 
    rename(
        female = sex_z_f,
        education_phd = education_ph_d,
        education_high_school = education_z_high_school,
        car_type_suv = car_type_z_suv,
        job_blue_collar = job_z_blue_collar,
        urbanicity_highly_rural = urbanicity_z_highly_rural_rural
    ) |> 
    select(-index)

insurance_eval_dummies <- insurance_eval |>
    mutate(
        across(all_of(currency_vars), parse_number)
    ) |>
    fastDummies::dummy_cols(
        select_columns = char_vars,
        remove_selected_columns = TRUE,
        remove_first_dummy = TRUE
    ) |> 
    clean_names()|> 
    rename(
        female = sex_z_f,
        education_phd = education_ph_d,
        education_high_school = education_z_high_school,
        car_type_suv = car_type_z_suv,
        job_blue_collar = job_z_blue_collar,
        urbanicity_highly_rural = urbanicity_z_highly_rural_rural
    ) |> 
    select(-index)

```

In previewing summary statistics for all variables, some of the minimum values appear to be in error. For example, `car_age` has a negative value, and there are a high number of 0 values in `car_age`, `home_val`, which may be data entry errors--we will replace them missing values.

```{r}
#| label: preview-summ-stats
#| output: false

insurance_train_dummies |> summary()

insurance_train_dummies <- insurance_train_dummies |>
    mutate(
        car_age = ifelse(car_age <= 0, NA, car_age),
        home_val = ifelse(home_val == 0, NA, home_val)
    )
```

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for all variables in the training dataset"
#| echo: false

tbl_summ_stats <- insurance_train_dummies |>
    skim_without_charts() |>
    janitor::clean_names() |>
    as_tibble() |>
    rename(Variable = skim_variable, "Completeness rate" = complete_rate, "Mean" = numeric_mean, "SD" = numeric_sd, "P50" = numeric_p50, "Min" = numeric_p0, "P25" = numeric_p25, "P75" = numeric_p75, "Max" = numeric_p100) |>
    select(Variable, "Completeness rate", Mean, Min, P25, P50, P75, Max) |>
    mutate(
        across(where(is.numeric), ~ signif(., 2)),
        Mean = format(Mean, scientific = FALSE, trim = TRUE)
    ) |>
    knitr::kable()

tbl_summ_stats
```

@tbl-summary-statistics presents descriptive characteristics of the training data. The target variables are `target_flag`, which represents whether the person was in a car crash (1) or not (0), and `target_amt`, which represents the cost of the crash if it occurs. The other variables represent various information on auto insurance for different individuals. 

```{r}
#| label: fig-correlation-matrix
#| fig-cap: "Correlation matrix of all variables in the training dataset. Note: Correlations with absolute value less than 0.2 are suppressed."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Get correlation matrix
insurance_train_cor <- insurance_train_dummies |>
    select(where(is.numeric)) |> 
    cor(use = "pairwise.complete.obs")

insurance_train_cor_plot <- insurance_train_cor
insurance_train_cor_plot[abs(insurance_train_cor_plot) < 0.2] <- NA

corrplot(insurance_train_cor_plot, method = "circle", type = "upper", sig.level = 0.05, insig = "blank", addCoef.col = "black", number.cex = 0.7, tl.cex = 0.7, tl.srt = 45, na.label = " ")

most_corr_w_target_flag <- insurance_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "target_flag", Variable != "target_amt") |>
    transmute(
        Variable,
        Correlation = target_flag,
        Abs_Correlation = abs(target_flag)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.2) |> pull(Variable)

most_corr_w_target_amt <- insurance_train_cor |>
    as.data.frame() |>
    rownames_to_column(var = "Variable") |>
    filter(Variable != "target_flag", Variable != "target_amt") |>
    transmute(
        Variable,
        Correlation = target_amt,
        Abs_Correlation = abs(target_amt)
    ) |>
    arrange(desc(Abs_Correlation)) |>
    filter(Abs_Correlation > 0.2) |> 
    pull(Variable)
```

The correlation between all variables are shown @fig-correlation-matrix. The variables most correlated with `target_flag`, whether there was a car crash, (besides the amount the crash costed) are `r paste(most_corr_w_target_flag, collapse = ", ")`. This means that people who work or live in a more rural area are less likely to be in a crash, and people who  have more motor vehicle record points or have more claims in the past 5 years are more likely to be in a crash. There were no predictor variables strongly correlated with `target_amt`, the cost if the crash occurs. 

Lastly, for continuous variables, we also looked at the distribution of each variable to check for skewness and outliers. 

```{r}
#| label: fig-distributions
#| fig-cap: "Distribution of continuous variables in the training dataset."
#| fig.align: "center"
#| fig-width: 10
#| fig-height: 10
#| warning: false

# Keep only continuous
continuous_vars <- insurance_train_dummies |>
    select(where(is.numeric)) |>
    select(where(~ dplyr::n_distinct(stats::na.omit(.x)) > 10)) |>
    names()

insurance_train_dummies |>
    select(all_of(continuous_vars)) |>
    pivot_longer(cols = everything(), names_to = "variable", values_to = "value") |>
    ggplot(aes(x = value)) +
    geom_histogram(bins = 30, fill = "steelblue", color = "white") +
    facet_wrap(~ variable, scales = "free", ncol = 4) +
    theme_minimal() +
    theme(panel.grid = element_blank()) +
    labs(x = NULL, y = "Count")

```

There are many 1-year old cars in the dataset, showing a high spike at 1 in the `car_age` variable. Dollar amounts such as `bluebook`, `home_val`, `income`, and `target_amt` are right-skewed, which is expected for financial variables. The same goes for travel time (`travtime`). Most people have no motor vehicle record points (`mvr_pts`) or claims in the past 5 years (`oldclaim`), which is expected for a general population. Years on the job (`yoj`) is almost normally distributed, but has a spike at 0, which may represent people who are unemployed or have been on the job for less than a year.

# Data Preparation

The goal of this section is to prepare the insurance training and evaluation data for modeling. The assignment requires transformations of the original variables or the creation of new variables when useful. In this dataset, preparation included missing value treatment, missingness flags, mathematical transformations, and derived variables.

```{r}
#| label: data-preparation
#| warning: false

# Identify numeric predictor variables
numeric_vars <- insurance_train_dummies |>
  select(where(is.numeric)) |>
  select(-target_flag, -target_amt) |>
  names()

# Create missingness flags before imputation
insurance_train_prepared <- insurance_train_dummies |>
  mutate(
    age_missing = ifelse(is.na(age), 1, 0),
    car_age_missing = ifelse(is.na(car_age), 1, 0),
    home_val_missing = ifelse(is.na(home_val), 1, 0),
    income_missing = ifelse(is.na(income), 1, 0),
    yoj_missing = ifelse(is.na(yoj), 1, 0)
  )

insurance_eval_prepared <- insurance_eval_dummies |>
  mutate(
    age_missing = ifelse(is.na(age), 1, 0),
    car_age_missing = ifelse(is.na(car_age), 1, 0),
    home_val_missing = ifelse(is.na(home_val), 1, 0),
    income_missing = ifelse(is.na(income), 1, 0),
    yoj_missing = ifelse(is.na(yoj), 1, 0)
  )

# Median imputation using training medians
for (v in numeric_vars) {
  med_val <- median(insurance_train_prepared[[v]], na.rm = TRUE)

  insurance_train_prepared[[v]][is.na(insurance_train_prepared[[v]])] <- med_val
  insurance_eval_prepared[[v]][is.na(insurance_eval_prepared[[v]])] <- med_val
}

# Create transformed and engineered variables
insurance_train_prepared <- insurance_train_prepared |>
  mutate(
    log_income = log1p(income),
    log_home_val = log1p(home_val),
    log_bluebook = log1p(bluebook),
    log_oldclaim = log1p(oldclaim),
    log_travtime = log1p(travtime),

    claim_rate = oldclaim / (clm_freq + 1),
    car_value_age_ratio = bluebook / (car_age + 1),
    risk_points_claims = mvr_pts + clm_freq,

    age_bucket = case_when(
      age < 25 ~ "young_driver",
      age >= 25 & age < 60 ~ "adult_driver",
      age >= 60 ~ "older_driver",
      TRUE ~ "unknown"
    )
  )

insurance_eval_prepared <- insurance_eval_prepared |>
  mutate(
    log_income = log1p(income),
    log_home_val = log1p(home_val),
    log_bluebook = log1p(bluebook),
    log_oldclaim = log1p(oldclaim),
    log_travtime = log1p(travtime),

    claim_rate = oldclaim / (clm_freq + 1),
    car_value_age_ratio = bluebook / (car_age + 1),
    risk_points_claims = mvr_pts + clm_freq,

    age_bucket = case_when(
      age < 25 ~ "young_driver",
      age >= 25 & age < 60 ~ "adult_driver",
      age >= 60 ~ "older_driver",
      TRUE ~ "unknown"
    )
  )

# Convert age bucket into dummy variables
insurance_train_prepared <- fastDummies::dummy_cols(
  insurance_train_prepared,
  select_columns = "age_bucket",
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

insurance_eval_prepared <- fastDummies::dummy_cols(
  insurance_eval_prepared,
  select_columns = "age_bucket",
  remove_selected_columns = TRUE,
  remove_first_dummy = TRUE
)

# Confirm no missing values remain
sum(is.na(insurance_train_prepared))
sum(is.na(insurance_eval_prepared))
```


### Linearity Assumption Summary

The results above indicate that the assumption of linearity is satisfied for all independent variables. However, several were flagged as being unlikely to have predictive value in a multivariate linear model.

## Data Splits

First, we will split our dataset into a training and testing set. We will use the training set to build our models and the testing set to evaluate their performance. The training set will be used to fit the models, while the testing set will be used to assess how well the models generalize to new data.

```{r}
#| label: split-data
#| eval: false

set.seed(19940211)
train_indices <- sample(nrow(crime_clean), size = 0.8 * nrow(crime_clean))
crime_train_split <- crime_clean[train_indices, ]
crime_test_split <- crime_clean[-train_indices, ]
```

# Check assumptions for binary logistic regression

1.  Binary Outcome

```{r}
#| label: assumption-binary-outcome
#| echo: true
#| eval: false


# Confirming that the target is binary
crime_clean |>
    count(target) |>
    mutate(prop = n / sum(n))
```

The response variable "target" is confirmed to be a binary, therefore it takes two values ( 0: crime rate below the median) and (1: crime rate above the median). In the 466 neighborhoods in training data set 237 (50.9%) fall in the class of low crime and 228 (49.1)% are in the high crime neighborhood. There is no class imbalance, due to the near equal split meaning there is unlikely to be bias towards one class.

2.  Independence of Observations

\vspace{0.5em}

Each record in the dataset represents a distinct neighborhood, and there is no repeated-measures or clustered structure in the data. Therefore, the independence of observations assumption is satisfied by design and requires no further adjustment.

3.  No Multicollinearity (VIF Check)

```{r}
#| label: assumption-multicollinearity
#| echo: true
#| warning: false
#| eval: false

library(car)

# Fit a reference logistic model with original predictors
model_ref <- glm(
    target ~ zn + indus + chas + nox + rm + age + dis + rad + tax + ptratio
        + lstat + medv,
    data = crime_clean, family = binomial
)

# Compute VIF
vif_vals <- vif(model_ref) |>
    as.data.frame() |>
    rownames_to_column("Variable") |>
    rename(VIF = 2) |>
    arrange(desc(VIF))

vif_vals |> knitr::kable(caption = "Variance Inflation Factors for Reference Model")
```

The VIF values is calculated for all predictors to check for multicollinearity. A VIF above 10 is considered to be severe, while between 5 and 10 should warrant caution. None of the variable exceed a threshold of 10, indicating multicollinearity is not present.The highest VIF belongs to medv (median home value, VIF = 8.12), followed by rm (average rooms per dwelling, VIF = 5.81). These two variables are moderate and do not need requrie any removal. The remaining predictors fall below 5 and are acceptable levels of multicollinearity. There was strong pairwise correlation between "rad" and "tax" in the exploration section, howevr it is confirmed that multicollinearity does not exist in the dataset.

4.  Linearity of Log-Odds (Box-Tidwell Test)

```{r}
#| label: assumption-linearity-log-odds
#| echo: true
#| warning: false
#| eval: false

# Box-Tidwell test: add interaction of each continuous predictor with its log
# to test whether the log-odds relationship is linear
continuous_vars <- c(
    "zn", "indus", "nox", "rm", "age", "dis", "rad", "tax",
    "ptratio", "lstat", "medv"
)

# Create interaction terms x * log(x) for each continuous predictor
# (use log1p to handle zeros)
crime_bt <- crime_clean |>
    mutate(across(all_of(continuous_vars), ~ . * log1p(.), .names = "{.col}_logx"))

bt_formula <- as.formula(paste(
    "target ~",
    paste(continuous_vars, collapse = " + "),
    "+",
    paste(paste0(continuous_vars, "_logx"), collapse = " + ")
))

model_bt <- glm(bt_formula, data = crime_bt, family = binomial)

# Extract p-values for the interaction terms only
bt_results <- summary(model_bt)$coefficients |>
    as.data.frame() |>
    rownames_to_column("Term") |>
    filter(str_detect(Term, "_logx")) |>
    mutate(
        Variable = str_remove(Term, "_logx"),
        Significant = `Pr(>|z|)` < 0.05
    ) |>
    select(Variable, Estimate, `Pr(>|z|)`, Significant) |>
    mutate(across(where(is.numeric), ~ signif(., 3)))

bt_results |> knitr::kable(caption = "Box-Tidwell Test for Linearity of Log-Odds")
```

The Box-Tidwell test was used to assess whether each continuous predictor has a linear relationship with the log-odds of the outcome. A p-value greater than 0.05 violate the assumption. There are three variables that were flagged as significant:rm (p = 0.003), dis (p \< 0.001), and tax (p \< 0.001). This indicates that these predictors do not have strict lienear relationship with log-odds of high crime. Log transformation were applied to these variables in the data preparation section in order to be used in the model building. The remaining eight predictors did not have any evidence of violating the linearity assumption.

# Build models

We will build models for two separate outcomes:

1. **TARGET_FLAG** (binary): Whether the person was in a car crash (1) or not (0)
2. **TARGET_AMT** (continuous): The amount of money the crash cost if it occurs

For each outcome, we compare three models as per the textbook methods:
- Model 1: Full model with all predictors
- Model 2: Reduced model (removing highly correlated or insignificant variables)
- Model 3: Backward stepwise selection model

## Model Building Functions

The following function takes a preprocessed dataset and builds all models:

```{r}
#| label: build-models-function
#| eval: false

#' Build Insurance Prediction Models
#' 
#' This function takes preprocessed training and test datasets and builds
#' multiple models for both binary classification (TARGET_FLAG) and
#' linear regression (TARGET_AMT).
#'
#' @param train_data Preprocessed training dataset (dataframe)
#' @param test_data Preprocessed test dataset (dataframe)
#' @param target_flag_col Name of binary target column (default: "TARGET_FLAG")
#' @param target_amt_col Name of continuous target column (default: "TARGET_AMT")
#' @return List containing all models and evaluation metrics
build_insurance_models <- function(train_data, test_data, 
                                    target_flag_col = "TARGET_FLAG",
                                    target_amt_col = "TARGET_AMT") {
  
  # ============================================================================
  # PART 1: BINARY LOGISTIC REGRESSION FOR TARGET_FLAG
  # ============================================================================
  cat("\n=== Building Binary Logistic Regression Models ===\n")
  
  # Get predictor columns (exclude targets and index)
  exclude_cols <- c("INDEX", target_flag_col, target_amt_col)
  predictor_cols <- names(train_data) |> 
    setdiff(exclude_cols) |>
    keep(~ is.numeric(train_data[[.x]]))
  
  # Build formula for full model
  flag_formula <- as.formula(paste(target_flag_col, "~", paste(predictor_cols, collapse = "+")))
  
  # Model 1: Full Logistic Model
  cat("Building Model 1: Full Logistic Model...\n")
  model1_logistic <- glm(flag_formula, data = train_data, family = binomial)
  
  # Model 2: Reduced Logistic Model (remove multicollinear variables)
  # Based on correlation analysis - keep only important predictors
  reduced_predictors <- c(
    "KIDSDRIV", "AGE", "HOMEKIDS", "YOJ", "INCOME", "HOME_VAL",
    "TRAVTIME", "BLUEBOOK", "TIF", "OLDCLAIM", "CLM_FREQ", "MVR_PTS", "CAR_AGE"
  )
  reduced_predictors <- reduced_predictors[reduced_predictors %in% names(train_data)]
  
  flag_formula_reduced <- as.formula(
    paste(target_flag_col, "~", paste(reduced_predictors, collapse = "+"))
  )
  
  cat("Building Model 2: Reduced Logistic Model...\n")
  model2_logistic <- glm(flag_formula_reduced, data = train_data, family = binomial)
  
  # Model 3: Backward Stepwise Selection
  cat("Building Model 3: Backward Stepwise Logistic Model...\n")
  model_full_logistic <- glm(flag_formula, data = train_data, family = binomial)
  model3_logistic <- step(model_full_logistic, direction = "backward", trace = 0)
  
  # ============================================================================
  # PART 2: LINEAR REGRESSION FOR TARGET_AMT
  # ============================================================================
  cat("\n=== Building Linear Regression Models ===\n")
  
  # Build formula for linear regression
  amt_formula <- as.formula(paste(target_amt_col, "~", paste(predictor_cols, collapse = "+")))
  
  # Model 1: Full Linear Model
  cat("Building Model 1: Full Linear Model...\n")
  model1_linear <- lm(amt_formula, data = train_data)
  
  # Model 2: Reduced Linear Model
  amt_formula_reduced <- as.formula(
    paste(target_amt_col, "~", paste(reduced_predictors, collapse = "+"))
  )
  
  cat("Building Model 2: Reduced Linear Model...\n")
  model2_linear <- lm(amt_formula_reduced, data = train_data)
  
  # Model 3: Backward Stepwise Selection
  cat("Building Model 3: Backward Stepwise Linear Model...\n")
  model_full_linear <- lm(amt_formula, data = train_data)
  model3_linear <- step(model_full_linear, direction = "backward", trace = 0)
  
  # ============================================================================
  # RETURN ALL MODELS
  # ============================================================================
  list(
    logistic = list(
      model1 = model1_logistic,
      model2 = model2_logistic,
      model3 = model3_logistic
    ),
    linear = list(
      model1 = model1_linear,
      model2 = model2_linear,
      model3 = model3_linear
    )
  )
}
```

Now let's use this function to build the models:

```{r}
#| label: build-models-usage
#| eval: false
#| warning: false

# Assuming the data has been preprocessed and split by another teammate
# The preprocessed data should have:
# - Numeric TARGET_FLAG (0/1)
# - Numeric TARGET_AMT
# - All predictor variables as numeric

# Example usage (uncomment when data is ready):
# all_models <- build_insurance_models(
#   train_data = insurance_train_processed,
#   test_data = insurance_test_processed
# )

# Store model names for reference
logistic_model_names <- c("Full Logistic Model", "Reduced Logistic Model", "Stepwise Logistic Model")
linear_model_names <- c("Full Linear Model", "Reduced Linear Model", "Stepwise Linear Model")

cat("Model building functions defined successfully.\n")
cat("To use: Pass preprocessed train/test dataframes to build_insurance_models()\n")
```

## Model 1. Full Logistic and Linear Models

As shown below, the most significant predictors for `TARGET_FLAG` (whether a crash occurred) include variables related to the driver's profile and history. The AIC for this logistic model is `r\ AIC(model1_logistic)`.

Similarly, for `TARGET_AMT` (cost of crash), the full linear model identifies significant predictors. The R-squared for this model is `r\ summary(model1_linear)$r.squared`.

```{r}
#| label: build-model1-insurance
#| eval: false

# Model 1: Full Logistic Model (TARGET_FLAG)
# This model uses all available numeric predictors
model1_logistic <- glm(
    formula = TARGET_FLAG ~ .,
    data = train_data |> select(-INDEX, -TARGET_AMT),
    family = binomial
)

cat("\n=== Full Logistic Model Summary ===\n")
summary(model1_logistic)

# Model 1: Full Linear Model (TARGET_AMT)
# This model uses all available numeric predictors
model1_linear <- lm(
    formula = TARGET_AMT ~ .,
    data = train_data |> select(-INDEX, -TARGET_FLAG)
)

cat("\n=== Full Linear Model Summary ===\n")
summary(model1_linear)
```

## Model 2. Reduced Models (Addressing Multicollinearity)

In the Data Exploration section, we identified several variables with high correlations. For the reduced models, we remove highly correlated predictors while retaining theoretically important ones.

The AIC for the reduced logistic model is `r\ AIC(model2_logistic)`, and the R-squared for the reduced linear model is `r\ summary(model2_linear)$r.squared`.

```{r}
#| label: build-model2-insurance
#| eval: false

# Model 2: Reduced Logistic Model
# Keep only important predictors based on correlation analysis
reduced_vars <- c("KIDSDRIV", "AGE", "HOMEKIDS", "YOJ", "INCOME", "HOME_VAL",
                  "TRAVTIME", "BLUEBOOK", "TIF", "OLDCLAIM", "CLM_FREQ", "MVR_PTS", "CAR_AGE")

model2_logistic <- glm(
    formula = as.formula(paste("TARGET_FLAG ~", paste(reduced_vars, collapse = "+"))),
    data = train_data,
    family = binomial
)

cat("\n=== Reduced Logistic Model Summary ===\n")
summary(model2_logistic)

# Model 2: Reduced Linear Model
model2_linear <- lm(
    formula = as.formula(paste("TARGET_AMT ~", paste(reduced_vars, collapse = "+"))),
    data = train_data
)

cat("\n=== Reduced Linear Model Summary ===\n")
summary(model2_linear)
```

## Model 3. Backward Stepwise Selection Models

For this model, we performed backward stepwise selection from the full model to identify the most important predictors.

The AIC for the stepwise logistic model is `r\ AIC(model3_logistic)`, and the R-squared for the stepwise linear model is `r\ summary(model3_linear)$r.squared`.

```{r}
#| label: build-model3-insurance
#| eval: false

# Model 3: Backward Stepwise Logistic Model
full_logistic <- glm(
    formula = TARGET_FLAG ~ .,
    data = train_data |> select(-INDEX, -TARGET_AMT),
    family = binomial
)

model3_logistic <- step(full_logistic, direction = "backward", trace = 0)

cat("\n=== Stepwise Logistic Model Summary ===\n")
summary(model3_logistic)

# Model 3: Backward Stepwise Linear Model
full_linear <- lm(
    formula = TARGET_AMT ~ .,
    data = train_data |> select(-INDEX, -TARGET_FLAG)
)

model3_linear <- step(full_linear, direction = "backward", trace = 0)

cat("\n=== Stepwise Linear Model Summary ===\n")
summary(model3_linear)
```

Of these three models, the stepwise models have the lowest AIC for logistic and highest R-squared for linear. In the next section, we will evaluate the performance of these models to determine which ones are best for predicting the insurance outcomes.

# Select Models

We evaluate both logistic and linear regression models using standard metrics from the textbook methods.

## Evaluation Metrics Function

```{r}
#| label: model-evaluation-functions
#| eval: false

#' Evaluate Logistic Regression Models
#' @param model A glm logistic model
#' @param model_name Character name of the model
#' @param data Test dataframe
#' @param target_col Name of target column
#' @return Tibble with evaluation metrics
evaluate_logistic_model <- function(model, model_name, data, target_col = "TARGET_FLAG") {
    probs <- predict(model, newdata = data, type = "response")
    preds <- factor(ifelse(probs > 0.5, 1, 0), levels = c(0, 1))
    actuals <- factor(data[[target_col]], levels = c(0, 1))
    
    cm <- confusionMatrix(preds, actuals, positive = "1")
    roc_obj <- roc(data[[target_col]], probs, quiet = TRUE)
    auc_val <- as.numeric(pROC::auc(roc_obj))
    
    tibble(
        Model = model_name,
        Deviance = model$deviance,
        Accuracy = cm$overall["Accuracy"],
        Error_Rate = 1 - cm$overall["Accuracy"],
        Precision = cm$byClass["Precision"],
        Sensitivity = cm$byClass["Sensitivity"],
        Specificity = cm$byClass["Specificity"],
        F1 = cm$byClass["F1"],
        AUC = auc_val,
        AIC = AIC(model),
        BIC = BIC(model)
    )
}

#' Evaluate Linear Regression Models
#' @param model An lm linear model
#' @param model_name Character name of the model
#' @param data Test dataframe
#' @param target_col Name of target column
#' @return Tibble with evaluation metrics
evaluate_linear_model <- function(model, model_name, data, target_col = "TARGET_AMT") {
    preds <- predict(model, newdata = data)
    actuals <- data[[target_col]]
    
    # Calculate metrics
    mse <- mean((actuals - preds)^2)
    rmse <- sqrt(mse)
    mae <- mean(abs(actuals - preds))
    r2 <- summary(model)$r.squared
    adj_r2 <- summary(model)$adj.r.squared
    
    tibble(
        Model = model_name,
        MSE = mse,
        RMSE = rmse,
        MAE = mae,
        R_Squared = r2,
        Adj_R_Squared = adj_r2,
        AIC = AIC(model),
        BIC = BIC(model)
    )
}

# Evaluate all logistic models
cat("=== Evaluating Logistic Regression Models ===\n")
logistic_models <- list(
    "Full Logistic Model" = model1_logistic,
    "Reduced Logistic Model" = model2_logistic,
    "Stepwise Logistic Model" = model3_logistic
)

logistic_comparison <- imap_dfr(logistic_models, ~ {
    evaluate_logistic_model(.x, .y, test_data)
}) |> arrange(AIC)

# Display logistic model comparison
logistic_comparison |> 
    knitr::kable(digits = 4, caption = "Logistic Model Comparison")

# Select best logistic model
best_logistic <- logistic_comparison |> slice(1)
cat("\nBest Logistic Model:", best_logistic$Model, "\n")

# Evaluate all linear models
cat("\n=== Evaluating Linear Regression Models ===\n")
linear_models <- list(
    "Full Linear Model" = model1_linear,
    "Reduced Linear Model" = model2_linear,
    "Stepwise Linear Model" = model3_linear
)

linear_comparison <- imap_dfr(linear_models, ~ {
    evaluate_linear_model(.x, .y, test_data)
}) |> arrange(desc(R_Squared))

# Display linear model comparison
linear_comparison |> 
    knitr::kable(digits = 4, caption = "Linear Model Comparison")

# Select best linear model
best_linear <- linear_comparison |> slice(1)
cat("\nBest Linear Model:", best_linear$Model, "\n")
```

For the logistic models, we use AIC as our primary criterion for model selection, with lower AIC indicating better model fit. The `r best_logistic$Model` achieved the lowest AIC of `r best_logistic$AIC`, with an accuracy of `r round(best_logistic$Accuracy, 3)` and AUC of `r round(best_logistic$AUC, 3)`.

For the linear regression models, we use R-squared to select the best model, with higher R-squared indicating better model fit. The `r best_linear$Model` achieved the highest R-squared of `r round(best_linear$R_Squared, 3)` with RMSE of `r round(best_linear$RMSE, 2)`.

## Selected Model Summaries

```{r}
#| label: selected-model-summaries
#| eval: false

# Display summary of best logistic model
cat("=== Best Logistic Model Summary ===\n")
summary(logistic_models[[best_logistic$Model]])

# Display summary of best linear model
cat("\n=== Best Linear Model Summary ===\n")
summary(linear_models[[best_linear$Model]])
```
```

To make inferences from our top performing models, we can look at the significance of the predictors.

For the logistic regression model (TARGET_FLAG), the model identifies key factors that significantly predict whether a person will be in a car crash. The most important predictors include variables related to driving history, vehicle characteristics, and demographic factors.

For the linear regression model (TARGET_AMT), significant predictors help explain the variation in the cost of claims when crashes occur.

# Results

Generate predictions on the evaluation dataset using the selected models:

```{r}
#| label: predictions-on-eval-set
#| echo: true
#| eval: false

# Get the best model names
best_logistic_name <- best_logistic$Model
best_linear_name <- best_linear$Model

# Generate predictions for TARGET_FLAG (probability of crash)
eval_probs <- predict(
    logistic_models[[best_logistic_name]], 
    newdata = eval_data, 
    type = "response"
)

# Generate predictions for TARGET_AMT (cost if crash)
eval_amounts <- predict(
    linear_models[[best_linear_name]], 
    newdata = eval_data
)

# Create results dataframe
insurance_eval_results <- eval_data |>
    mutate(
        PREDICTED_TARGET_FLAG = ifelse(eval_probs > 0.5, 1, 0),
        PREDICTED_TARGET_AMT = ifelse(PREDICTED_TARGET_FLAG == 1, 
                                       round(eval_amounts, 2), 
                                       0)
    )

# Write predictions to CSV
write_csv(insurance_eval_results, "insurance_evaluation_predictions.csv")

# Display sample predictions
cat("Sample predictions for evaluation data:\n")
insurance_eval_results |>
    select(INDEX, PREDICTED_TARGET_FLAG, PREDICTED_TARGET_AMT) |>
    head(10) |>
    knitr::kable(caption = "Sample Insurance Predictions")
```

We generate predictions for both outcomes:

1. **TARGET_FLAG**: Whether the person will be in a car crash (binary: 0 or 1)
2. **TARGET_AMT**: The cost if a crash occurs (continuous dollar amount)

The predictions are written to `insurance_evaluation_predictions.csv` for submission.

# Conclusion

In conclusion, our `best_model_info$Model` model, which was selected based on deviance and other performance metrics, demonstrates strong predictive capabilities for classifying neighborhoods as high-crime or low-crime based on a variety of environmental and structural factors. The significant predictors identified in the model provide valuable insights into the underlying drivers of crime in urban areas, highlighting the importance of air pollution, school crowding, aging infrastructure, and proximity to employment centers. These findings can inform policymakers and community leaders in their efforts to target interventions and allocate resources effectively to reduce crime rates. Future work could explore additional modeling techniques, such as regularization or ensemble methods, to further enhance predictive performance and interpretability.
